---
title: "STAT 812/420"
subtitle: "Random Numbers and Elementary Monte Carlo"
author: "Longhai Li"
format: banner_slides-revealjs
execute:
  echo: true
  fig-width: 7
  fig-height: 4
---

```{r}
#| echo: false
set.seed(812)
```

# Generating Random Numbers

## Pseudo Random Numbers

A computer is deterministic, so it cannot produce truly random numbers. It produces **pseudo random numbers**: a deterministic sequence that *behaves* like iid $\mathrm{Unif}(0,1)$ draws.

A classic generator, the **linear congruential generator (LCG)**:
$$x_{i+1} = (a\,x_i + c) \bmod m, \qquad u_{i+1} = x_{i+1}/m$$

- The starting value $x_0$ is the **seed**. The same seed gives the same sequence, which makes simulations reproducible.
- The sequence eventually repeats; a good generator has a very long period and passes statistical tests of uniformity and independence.
- R's default is the Mersenne Twister (period $2^{19937}-1$). All other distributions in R are built from $\mathrm{Unif}(0,1)$ draws.

```{r}
lcg <- function(n, seed, a = 1103515245, c = 12345, m = 2^31) {
  x <- numeric(n); x[1] <- seed
  for (i in 2:n) x[i] <- (a * x[i - 1] + c) %% m
  x / m
}
round(lcg(5, seed = 812), 4)
```

## Seeds and Reproducibility

```{r}
set.seed(1); runif(3)
set.seed(1); runif(3)      # identical
runif(3)                   # continues the sequence
```

Always `set.seed()` at the top of a simulation study so results can be reproduced and debugged.

## Example: Generating $\mathrm{Unif}(a,b)$

If $U \sim \mathrm{Unif}(0,1)$ then $X = a + (b-a)\,U \sim \mathrm{Unif}(a,b)$:
$$P(X \le x) = P\!\left(U \le \frac{x-a}{b-a}\right) = \frac{x-a}{b-a}, \qquad a \le x \le b .$$

```{r}
u <- runif(10000); x <- 2 + 3 * u          # Unif(2, 5)
hist(x, breaks = 30, freq = FALSE, main = "a + (b - a) U with a = 2, b = 5")
abline(h = 1/3, col = 2, lwd = 2)
```

## Inverting the CDF (Continuous Case)

Let $F_X$ be a continuous CDF with inverse (quantile function) $F_X^{-1}$.

**Theorem.** If $U \sim \mathrm{Unif}(0,1)$, then $X = F_X^{-1}(U)$ has CDF $F_X$.

**Proof.**
$$P(X \le x) = P\big(F_X^{-1}(U) \le x\big) = P\big(U \le F_X(x)\big) = F_X(x).$$

Equivalently, by the change-of-variable formula with $u = F_X(x)$:
$$f_X(x) = f_U\big(F_X(x)\big)\left|\frac{du}{dx}\right| = 1 \times f(x).$$

The converse also holds: if $X \sim F_X$ then $F_X(X) \sim \mathrm{Unif}(0,1)$ (the **probability integral transform**), which we will use for p-values.

## Inverting the CDF: Picture

```{r}
#| echo: false
curve(pexp(x, 1), 0, 5, lwd = 2, xlab = "x", ylab = "F(x)", main = "X = F^{-1}(U)")
u <- c(0.2, 0.6, 0.9); x <- qexp(u, 1)
segments(0, u, x, u, col = 2, lty = 2); segments(x, 0, x, u, col = 4, lty = 2)
points(x, u, pch = 19); axis(2, u, labels = paste0("u=", u), las = 1, cex.axis = 0.8)
axis(1, x, labels = round(x, 2), cex.axis = 0.8)
```

Draw $u$ on the vertical axis, read $x = F^{-1}(u)$ off the horizontal axis. Regions where $F$ is steep (high density) receive more draws.

## Example: Exponential Distribution

$X \sim \mathrm{Exp}(\lambda)$ has $F_X(x) = 1 - e^{-\lambda x}$ for $x > 0$. Solving $u = 1 - e^{-\lambda x}$:
$$x = F_X^{-1}(u) = -\frac{1}{\lambda}\log(1-u).$$

Since $1 - U \sim \mathrm{Unif}(0,1)$ as well, we may use $X = -\log(U)/\lambda$.

```{r}
lambda <- 2; u <- runif(10000); x <- -log(u) / lambda
hist(x, breaks = 40, freq = FALSE, main = "Exp(2) via inverse CDF")
curve(dexp(x, lambda), add = TRUE, col = 2, lwd = 2)
c(sample_mean = mean(x), true_mean = 1 / lambda)
```

## Inverting the CDF (Discrete Case)

For a discrete $X$ with values $x_1 < x_2 < \cdots$ and CDF $F$, set
$$X = x_k \quad\text{where } F(x_{k-1}) < U \le F(x_k).$$

That is, cut $(0,1)$ into intervals of lengths $P(X = x_k)$ and see which one $U$ lands in.

```{r}
p <- c(0.2, 0.5, 0.3); vals <- c(1, 2, 3)
u <- runif(10000)
x <- vals[findInterval(u, cumsum(p)) + 1]   # or sample(vals, n, TRUE, p)
table(x) / length(x)
```

## A Special Transformation for Generating Normal Samples

$\Phi^{-1}$ has no closed form, so the inverse-CDF method is not directly usable. Instead, look at a pair $(Z_1, Z_2) \overset{iid}{\sim} N(0,1)$ in **polar coordinates**:
$$Z_1 = R\cos\Theta, \qquad Z_2 = R\sin\Theta, \qquad R^2 = Z_1^2 + Z_2^2, \quad \Theta = \operatorname{atan2}(Z_2, Z_1).$$

The joint density $\frac{1}{2\pi}e^{-(z_1^2+z_2^2)/2}$ depends on $(z_1, z_2)$ only through $r^2 = z_1^2 + z_2^2$, so it is **rotationally symmetric** about the origin. Consequently:

- The angle $\Theta \sim \mathrm{Unif}(0, 2\pi)$, independent of $R$.
- The squared radius $R^2 = Z_1^2 + Z_2^2 \sim \chi^2_2$, and $\chi^2_2$ **is** the exponential distribution with rate $1/2$:
$$f_{\chi^2_2}(w) = \frac{1}{2}e^{-w/2}, \quad w > 0 \qquad\Longrightarrow\qquad R^2 \sim \mathrm{Exp}(\text{rate} = 1/2).$$

Both $R^2$ and $\Theta$ **can** be generated by inverting a CDF.

## Illustration: Two Standard Normals in Polar Coordinates

```{r}
#| echo: false
library(latex2exp)
z1 <- rnorm(3000); z2 <- rnorm(3000); r2 <- z1^2 + z2^2; th <- atan2(z2, z1) %% (2 * pi)
layout(matrix(c(1, 1, 2, 3), 2, 2), widths = c(1.1, 1)); par(pty = "s", mar = c(4, 4, 2, 1))
plot(z1, z2, pch = 20, cex = 0.4, col = "grey40", asp = 1, xlab = TeX("$Z_1$"), ylab = TeX("$Z_2$"),
     main = TeX("$(Z_1, Z_2)$ iid N(0,1)"))
q <- sqrt(qchisq(c(0.5, 0.9, 0.99), 2))
for (r in q) symbols(0, 0, circles = r, inches = FALSE, add = TRUE, fg = 2, lwd = 2)
text(q + 0.25, 0.25, c("50%", "90%", "99%"), col = 2, cex = 0.8)
par(pty = "m")
hist(r2, breaks = 40, freq = FALSE, main = TeX("$R^2 = Z_1^2 + Z_2^2$"), xlab = "")
curve(dexp(x, 1/2), add = TRUE, col = 2, lwd = 2); legend("topright", "Exp(rate 1/2)", col = 2, lwd = 2, bty = "n")
hist(th, breaks = 24, freq = FALSE, main = TeX("$\\Theta$"), xlab = ""); abline(h = 1 / (2 * pi), col = 2, lwd = 2)
```

The point cloud is rotationally symmetric: the red circles enclose 50%, 90%, 99% of the points ($R^2 \le \chi^2_2$ quantiles); the angle is uniform; $R^2$ follows $\mathrm{Exp}(1/2)$.

## The Box–Muller Transformation

Reverse the picture: **generate** $R^2$ and $\Theta$, then rotate back to $(Z_1, Z_2)$.

1. $R^2 \sim \mathrm{Exp}(1/2)$ by inverse CDF: $F(w) = 1 - e^{-w/2}$, so $R^2 = -2\log U_1$, i.e. $R = \sqrt{-2\log U_1}$.
2. $\Theta \sim \mathrm{Unif}(0, 2\pi)$: $\Theta = 2\pi U_2$.
3. $Z_1 = R\cos\Theta$, $Z_2 = R\sin\Theta$.

**Theorem (Box–Muller).** If $U_1, U_2 \overset{iid}{\sim} \mathrm{Unif}(0,1)$, then
$$Z_1 = \sqrt{-2\log U_1}\,\cos(2\pi U_2), \qquad Z_2 = \sqrt{-2\log U_1}\,\sin(2\pi U_2)$$
are independent $N(0,1)$. Two uniforms produce two normals. Then $\mu + \sigma Z \sim N(\mu, \sigma^2)$.

## Box–Muller in R

```{r}
box_muller <- function(n) {
  u1 <- runif(n); u2 <- runif(n)
  r  <- sqrt(-2 * log(u1))            # R^2 ~ Exp(1/2), i.e. chi-square(2)
  th <- 2 * pi * u2                   # Theta ~ Unif(0, 2 pi)
  cbind(z1 = r * cos(th), z2 = r * sin(th))
}
z <- box_muller(3000)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
par(pty = "s"); plot(z, pch = 20, cex = 0.4, col = "grey40", asp = 1, main = "Box-Muller output")
par(pty = "m"); qqnorm(c(z), pch = ".", main = "Normal Q-Q plot of both coordinates"); qqline(c(z), col = 2)
c(mean = mean(z), sd = sd(z), corr = cor(z)[1, 2])
```

# Monte Carlo Integration

## Why Monte Carlo Works: Two Theorems

Monte Carlo estimates an expectation $\mu = E(g(X))$ by an average of simulated values,
$$\hat\mu_n = \frac{1}{n}\sum_{i=1}^n g(X_i), \qquad X_i \overset{iid}{\sim} f .$$

**Theorem 5.1.1 (Strong Law of Large Numbers).** If $X_1, X_2, \ldots$ are iid with finite mean $\mu$, then
$$P\left(\frac{X_1 + \cdots + X_n}{n} \to \mu\right) = 1 . \tag{5.1}$$

The LLN guarantees that the average **converges** to the truth. It does not say how fast.

**Theorem 5.1.2 (Central Limit Theorem).** If in addition the variance $\sigma^2$ is finite, then
$$\frac{\bar X - \mu}{\sigma/\sqrt n} \;\xrightarrow{d}\; N(0,1). \tag{5.2}$$

The CLT gives the **error bound**: the Monte Carlo error is of order $\sigma/\sqrt n$.

## Monte Carlo Error Bound from the CLT

From (5.2), with confidence level 95%,
$$P\!\left(|\bar X - \mu| \le 1.96\,\frac{\sigma}{\sqrt n}\right) \approx 0.95, \tag{5.3}$$
so a 95% interval for $\mu$ is
$$\bar X \pm 1.96\,\frac{\sigma}{\sqrt n}, \qquad \sigma \text{ replaced by the sample SD } s .$$

- Halving the error requires **four times** as many samples; each extra decimal digit of accuracy costs a **100-fold** increase in $n$.
- The rate $n^{-1/2}$ does not depend on the dimension of $X$, which is why Monte Carlo beats numerical quadrature in high dimensions.

## Demonstration of the LLN

Running averages of $X_i \sim \mathrm{Exp}(1)$ ($\mu = 1$) for three independent sequences:

```{r}
n <- 5000
plot(NULL, xlim = c(1, n), ylim = c(0.5, 1.6), log = "x",
     xlab = "n (log scale)", ylab = "running average", main = "LLN: Exp(1)")
for (k in 1:3) lines(cumsum(rexp(n)) / (1:n), col = k + 1)
abline(h = 1, lty = 2)
```

## Demonstration of the CLT

For each of $N = 5000$ repetitions, draw $n$ values from $\mathrm{Exp}(1)$ and standardize the mean, $(\bar X - 1)/(1/\sqrt n)$:

```{r}
clt_demo <- function(n, N = 5000) replicate(N, (mean(rexp(n)) - 1) * sqrt(n))
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (n in c(2, 10, 100)) {
  hist(clt_demo(n), breaks = 40, freq = FALSE, xlim = c(-4, 4), main = paste("n =", n), xlab = "")
  curve(dnorm, add = TRUE, col = 2, lwd = 2)
}
```

The skewed exponential becomes approximately normal as $n$ grows.

## An Example: Estimating $\pi$ by Monte Carlo

Let $(X, Y)$ be uniform on the square $S = (-1,1)\times(-1,1)$, and let $C = \{(x,y): x^2 + y^2 \le 1\}$ be the unit disk. Then
$$P\big((X,Y) \in C\big) = \frac{\text{area}(C)}{\text{area}(S)} = \frac{\pi}{4}.$$

Define $Z = 4\, I\big((X,Y)\in C\big)$, so $E(Z) = \pi$ and
$$\mathrm{Var}(Z) = 16 \cdot \frac{\pi}{4}\left(1 - \frac{\pi}{4}\right) \approx 2.70, \qquad \mathrm{SD}(Z) \approx 1.64 .$$

Drawing $(X_i, Y_i)$ and averaging $Z_i$:
$$\hat\pi_n = \frac{1}{n}\sum_{i=1}^n Z_i = 4 \times \frac{\#\{i: X_i^2 + Y_i^2 \le 1\}}{n}. \tag{5.4}$$

## Visualizing the Method

```{r}
n <- 2000; x <- runif(n, -1, 1); y <- runif(n, -1, 1); inside <- x^2 + y^2 <= 1
par(pty = "s")
plot(x, y, col = ifelse(inside, "steelblue", "grey60"), pch = 20, cex = 0.6,
     main = sprintf("n = %d, proportion inside = %.3f, pi-hat = %.3f", n, mean(inside), 4 * mean(inside)))
curve(sqrt(1 - x^2), -1, 1, add = TRUE, lwd = 2); curve(-sqrt(1 - x^2), -1, 1, add = TRUE, lwd = 2)
```

## The Monte Carlo Estimate and Its Error

```{r}
est_pi <- function(n) {
  z <- 4 * (runif(n, -1, 1)^2 + runif(n, -1, 1)^2 <= 1)
  c(estimate = mean(z), se = sd(z) / sqrt(n))
}
r <- est_pi(10000)
r
r["estimate"] + c(-1.96, 1.96) * r["se"]    # 95% interval; contains pi = 3.14159?
```

The standard error $s/\sqrt n$ is computed from the **same** simulated values — no extra cost. Always report it with a Monte Carlo estimate.

## Results by Sample Size

```{r}
ns <- 10^(2:6)
res <- t(sapply(ns, est_pi))
data.frame(n = ns, estimate = round(res[, 1], 5), error = round(res[, 1] - pi, 5),
           se = signif(res[, 2], 3), theory_se = signif(1.64 / sqrt(ns), 3))
```

- The observed error is typically within about $2$ standard errors.
- SE shrinks by $\sqrt{10} \approx 3.2$ for each 10-fold increase in $n$, as the CLT predicts.

## Convergence of $\hat\pi_n$

```{r}
n <- 1e5; z <- 4 * (runif(n, -1, 1)^2 + runif(n, -1, 1)^2 <= 1)
k <- 1:n; pihat <- cumsum(z) / k
plot(k, pihat, type = "l", log = "x", ylim = c(2.8, 3.5), xlab = "n (log scale)", ylab = expression(hat(pi)[n]))
lines(k, pi + 1.96 * 1.64 / sqrt(k), lty = 2, col = 2); lines(k, pi - 1.96 * 1.64 / sqrt(k), lty = 2, col = 2)
abline(h = pi, col = 4)
```

The red band is $\pi \pm 1.96\,\sigma/\sqrt n$; the running estimate stays inside it most of the time.

# Evaluating Statistical Methods by Simulation

## Evaluating a Point Estimator

A point estimator $\hat\theta = \hat\theta(X_1, \ldots, X_n)$ is judged by its **mean squared error**
$$\mathrm{MSE}(\hat\theta) = E\big[(\hat\theta - \theta)^2\big] = \mathrm{Var}(\hat\theta) + \big[\mathrm{Bias}(\hat\theta)\big]^2 .$$

MSE is an expectation, so it is estimated by Monte Carlo: simulate $N$ independent data sets from a model with **known** $\theta$,
$$\hat\theta^{(j)} = \hat\theta\big(X_1^{(j)}, \ldots, X_n^{(j)}\big), \qquad j = 1, \ldots, N,$$
and compute
$$\widehat{\mathrm{MSE}} = \frac{1}{N}\sum_{j=1}^N \big(\hat\theta^{(j)} - \theta\big)^2, \qquad
\widehat{\mathrm{Bias}} = \bar{\hat\theta} - \theta, \qquad
\widehat{\mathrm{Var}} = \frac{1}{N}\sum_{j=1}^N \big(\hat\theta^{(j)} - \bar{\hat\theta}\big)^2 .$$

Two sample sizes appear: $n$ (data per replicate, fixed by the problem) and $N$ (replicates, chosen large enough that the Monte Carlo error is negligible).

## Example: Mean vs. Median vs. Trimmed Mean

Estimating the centre $\theta = 0$ of a symmetric distribution, $n = 20$, $N = 5000$ replicates.

```{r}
eval_est <- function(rdist, n = 20, N = 5000) {
  est <- replicate(N, { x <- rdist(n); c(mean = mean(x), median = median(x), trim10 = mean(x, trim = 0.1)) })
  rbind(bias = rowMeans(est), var = apply(est, 1, var), mse = rowMeans(est^2))
}
round(eval_est(rnorm), 4)                          # normal data
round(eval_est(function(n) rt(n, df = 3)), 4)      # heavy-tailed t3 data
```

No estimator is best everywhere: the mean wins under normality, the trimmed mean and median win under heavy tails.

# Evaluating and Conducting Hypothesis Tests

## Hypothesis Tests and the Two Types of Error

A test rejects $H_0$ when a statistic $T$ exceeds a critical value $c_\alpha$.

|          | reject: $T > c_\alpha$      | do not reject: $T \le c_\alpha$      |
|----------|---------------------|-----------------------|
| $H_0$ true    | Type I error (false positive)   | true negative                    |
| $H_1$ true    | true positive                  | Type II error (false negative)    |

$$\text{size} = P(T > c_\alpha \mid H_0) = \alpha, \qquad \text{power} = P(T > c_\alpha \mid H_1) = 1 - \beta .$$

Both are probabilities — hence expectations of indicators — so both can be **estimated by Monte Carlo** once we can simulate data under $H_0$ and $H_1$.

## Estimating the Critical Value $c_\alpha$ by Monte Carlo

Simulate $N$ data sets **under $H_0$** and compute $T^{(1)}, \ldots, T^{(N)}$:

| replicate | $T^{(j)}$ | $I(T^{(j)} > c_\alpha)$ |
|:---:|:---:|:---:|
| 1 | $T^{(1)}$ | 1 |
| 2 | $T^{(2)}$ | 0 |
| $\vdots$ | $\vdots$ | $\vdots$ |
| $N$ | $T^{(N)}$ | 0 |

$$\hat P(T > c_\alpha \mid H_0) = \frac{1}{N}\sum_{j=1}^N I\big(T^{(j)} > c_\alpha\big) \;\doteq\; \alpha$$

so take $\hat c_\alpha$ = the $(1-\alpha)$ sample quantile of $T^{(1)}, \ldots, T^{(N)}$. This replaces the (often unknown) null distribution of $T$ by its **empirical** distribution.

## P-value

The **p-value** is the probability, under $H_0$, of a statistic at least as extreme as the observed one:
$$pv(t) = P(T \ge t \mid H_0) = 1 - F_{H_0}(t),$$
the **survival function** of $T$ under $H_0$ evaluated at $t^{obs}$. Rejecting when $pv(t^{obs}) < \alpha$ is the same as rejecting when $t^{obs} > c_\alpha$.

```{r}
#| echo: false
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
t0 <- 1.8; curve(dchisq(x, 3), 0, 12, lwd = 2, main = "density of T | H0", xlab = "t", ylab = "")
xx <- seq(t0, 12, length = 100); polygon(c(t0, xx, 12), c(0, dchisq(xx, 3), 0), col = "lightblue", border = NA)
abline(v = t0, col = 2); text(t0 + 3, 0.15, "pv = shaded area")
curve(1 - pchisq(x, 3), 0, 12, lwd = 2, main = "survival function 1 - F(t)", xlab = "t", ylab = "")
abline(v = t0, col = 2); abline(h = 1 - pchisq(t0, 3), lty = 2)
```

## Uniformity of $pv(T)$ under $H_0$

**Theorem.** If $T \mid H_0$ has continuous CDF $F_{H_0}$, then $pv(T) = 1 - F_{H_0}(T) \sim \mathrm{Unif}(0,1)$ under $H_0$.

**Proof.** For $t \in (0,1)$,
$$
\begin{aligned}
P\big(pv(T) < t\big) &= P\big(1 - F_{H_0}(T) < t\big) = P\big(F_{H_0}(T) > 1 - t\big) \\
&= P\big(T > F_{H_0}^{-1}(1-t)\big) = 1 - F_{H_0}\big(F_{H_0}^{-1}(1-t)\big) = t .
\end{aligned}
$$

Consequences:

- $P(pv < \alpha \mid H_0) = \alpha$ **exactly**: rejecting when $pv < \alpha$ has size $\alpha$.
- A histogram of simulated p-values under $H_0$ should be flat — the standard **test of a test**.

## P-value Distribution under $H_0$ vs. $H_1$

One-sample $t$-test of $H_0: \mu = 0$, $n = 20$, $N = 5000$:

```{r}
pv_sim <- function(mu, n = 20, N = 5000) replicate(N, t.test(rnorm(n, mu))$p.value)
pv0 <- pv_sim(0); pv1 <- pv_sim(0.5)
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
hist(pv0, breaks = 20, main = expression(H[0]: mu == 0), xlab = "p-value"); abline(h = 250, col = 2, lty = 2)
hist(pv1, breaks = 20, main = expression(H[1]: mu == 0.5), xlab = "p-value")
```

Under $H_0$ the p-values are uniform; under $H_1$ they pile up near 0.

## Monte Carlo Estimates of Size and Power

$$\widehat{\text{size}} = \hat P(pv < \alpha \mid H_0) = \frac{1}{N}\sum_{j=1}^N I\big(pv^{(j)}_{H_0} < \alpha\big), \qquad
\widehat{\text{power}} = \hat P(pv < \alpha \mid H_1) = \frac{1}{N}\sum_{j=1}^N I\big(pv^{(j)}_{H_1} < \alpha\big)$$

```{r}
c(size = mean(pv0 < 0.05), power = mean(pv1 < 0.05),
  exact_power = power.t.test(n = 20, delta = 0.5, sd = 1)$power)
```

Each estimate is a proportion, so its Monte Carlo standard error is $\sqrt{\hat p(1-\hat p)/N}$ — about $0.003$ for the size with $N = 5000$.

## Monte Carlo for Calculating a P-value

When $F_{H_0}$ is unknown, estimate $pv(t^{obs})$ by simulating $T$ under $H_0$:
$$\widehat{pv}(t^{obs}) = \frac{1 + \sum_{j=1}^N I\big(T^{(j)} \ge t^{obs}\big)}{N + 1}.$$

(The "+1" counts the observed statistic as one draw from $H_0$; it keeps the estimate above 0 and the test exactly valid.)

Example: test $H_0: \mu = 0$ for $n = 15$ observations using $T = |\bar X|/(s/\sqrt n)$, but suppose we did not know the $t$-distribution.

```{r}
x <- rnorm(15, mean = 0.6); tobs <- abs(mean(x)) / (sd(x) / sqrt(15))
Tnull <- replicate(5000, { y <- rnorm(15); abs(mean(y)) / (sd(y) / sqrt(15)) })
c(mc_pv = (1 + sum(Tnull >= tobs)) / 5001, exact_pv = 2 * pt(-tobs, df = 14))
```

## A Permutation Test for the Correlation Coefficient

Data $(X_i, Y_i)$, $i = 1, \ldots, n$. Test $H_0: X \perp Y$ (which implies $\mathrm{corr}(X,Y) = 0$) with $T = |\hat\rho|$.

Under $H_0$, **every pairing of the $X$ values with the $Y$ values is equally likely**, so the null distribution of $T$ is obtained by re-computing $T$ after randomly permuting $Y$:

1. Compute $\rho^{obs} = \hat\rho(X, Y)$.
2. For $j = 1, \ldots, N$: permute $Y$ to get $Y^{(j)}$, compute $\rho^{(j)} = \hat\rho(X, Y^{(j)})$.
3. $\widehat{pv} = \dfrac{1 + \#\{j : |\rho^{(j)}| \ge |\rho^{obs}|\}}{N+1}$.

No normality assumption is needed; the null distribution is generated **from the data itself**.

## Permutation Test for Correlation in R

```{r}
perm_cor_test <- function(x, y, N = 2000) {
  robs <- cor(x, y)
  rperm <- replicate(N, cor(x, sample(y)))
  list(rho = robs, pv = (1 + sum(abs(rperm) >= abs(robs))) / (N + 1), rperm = rperm)
}
x <- rexp(30); y <- 0.4 * x + rexp(30)     # non-normal, dependent
out <- perm_cor_test(x, y)
hist(out$rperm, breaks = 40, main = "permutation distribution of rho", xlab = "rho")
abline(v = c(-1, 1) * abs(out$rho), col = 2, lwd = 2)
c(rho_obs = out$rho, perm_pv = out$pv, pearson_pv = cor.test(x, y)$p.value)
```

## Example: Testing a Regression With Variable Selection

**Setting.** Response $Y$ and $p = 20$ candidate predictors $X_1, \ldots, X_{20}$, $n = 50$. We want to test $H_0$: $Y$ is unrelated to **all** of the $X$'s.

A common practice:

1. Pick the predictor $X_{k^*}$ most correlated with $Y$.
2. Fit $Y \sim X_{k^*}$ and report the usual $t$-test p-value for its slope.

**Question.** Is this a valid test of $H_0$? Does it have size $\alpha$?

## A Naive (Incorrect) Test

```{r}
naive_test <- function(y, X) {
  k <- which.max(abs(cor(X, y)))
  summary(lm(y ~ X[, k]))$coefficients[2, 4]    # t-test p-value of selected slope
}
```

The $t$-test p-value assumes $X_{k^*}$ was **chosen in advance**. Selecting the best of 20 predictors *after* looking at the data means that, even under $H_0$, $\max_k |\hat\rho_k|$ tends to be large. The reported p-value ignores this selection.

## Simulation for Evaluating the Naive Test

Generate data under $H_0$ ($Y$ independent of all $X$'s) and check whether the p-values are uniform:

```{r}
sim_null <- function(n = 50, p = 20) list(y = rnorm(n), X = matrix(rnorm(n * p), n, p))
pv_naive <- replicate(2000, { d <- sim_null(); naive_test(d$y, d$X) })
hist(pv_naive, breaks = 20, main = "naive test p-values under H0", xlab = "p-value")
c(size_at_0.05 = mean(pv_naive < 0.05))
```

The true size is far above 0.05: the naive test rejects a true $H_0$ far too often. Note that under $H_0$ the naive p-value is **not** uniform, which is a general diagnostic for an invalid test.

## Permutation Test (Full Procedure)

Fix: treat the **entire procedure** — selection plus fitting — as the test statistic, and obtain its null distribution by permuting $Y$.

$$T(Y, X) = \max_k |\hat\rho(X_k, Y)|$$

1. Compute $T^{obs} = T(Y, X)$.
2. For $j = 1, \ldots, N$: permute $Y$, compute $T^{(j)} = T(Y^{(j)}, X)$ (**re-doing the selection** each time).
3. $\widehat{pv} = \dfrac{1 + \#\{j : T^{(j)} \ge T^{obs}\}}{N+1}$.

```{r}
perm_test <- function(y, X, N = 500) {
  Tstat <- function(y) max(abs(cor(X, y)))
  Tobs <- Tstat(y); Tperm <- replicate(N, Tstat(sample(y)))
  (1 + sum(Tperm >= Tobs)) / (N + 1)
}
```

## Size and Power of the Permutation Test

```{r}
#| cache: true
sim_alt <- function(n = 50, p = 20, beta = 0.5) {
  X <- matrix(rnorm(n * p), n, p); list(y = beta * X[, 1] + rnorm(n), X = X)
}
pv_perm0 <- replicate(300, { d <- sim_null(); perm_test(d$y, d$X) })
pv_perm1 <- replicate(300, { d <- sim_alt();  perm_test(d$y, d$X) })
pv_naive1 <- replicate(300, { d <- sim_alt(); naive_test(d$y, d$X) })
rbind(naive = c(size = mean(pv_naive < 0.05), power = mean(pv_naive1 < 0.05)),
      permutation = c(size = mean(pv_perm0 < 0.05), power = mean(pv_perm1 < 0.05)))
```

- The permutation test has the correct size; its power is honest, whereas the naive test's "power" is inflated by its false positives.
- Power comparisons are only meaningful **between tests of the same size**.

## Summary

- All simulation starts from pseudo-random $\mathrm{Unif}(0,1)$ draws; other distributions come from **inverse-CDF** or **transformations** (Box–Muller).
- The LLN justifies Monte Carlo averages; the CLT gives their error, $\sigma/\sqrt n$, which is reported as a standard error.
- Bias, variance and MSE of estimators, and size and power of tests, are expectations — simulate with known truth and average.
- P-values are uniform under $H_0$: a flat histogram of simulated p-values is the basic test of a test.
- When the null distribution is unknown, generate it by simulation or by **permutation**; the permuted statistic must reproduce the **entire** data-analysis procedure.
