---
subtitle: "Computer Arithmetic"
format: banner_slides-revealjs
code-fold: show
execute:
  echo: true
include-in-header:
  text: |
    <style>
    .pv table { border-collapse: collapse; margin: 0.4em auto; font-family: monospace; }
    .pv td, .pv th { border: 2px solid #555; min-width: 2em; height: 1.5em;
                     text-align: center !important; padding: 0.3em 0.3em; font-size: 0.8em; }
    .pv th { border: none; font-size: 0.55em; color: #666; font-weight: normal; }
    .pv td.s { background: #fbe4ec; }   /* sign */
    .pv td.m { background: #e3eefb; }   /* mantissa / digits */
    .pv td.e { background: #e7f6e0; }   /* exponent */
    .pv td.x { background: #ffd6d6; color: #b00000; font-weight: bold; } /* lost / invalid */
    .pv td.p { border: none; padding: 0; min-width: 0; font-weight: bold; } /* punctuation */
    .pv td.z { color: #bbb; border-style: dashed; }     /* alignment zero, not stored */
    .pv td.op { border: none; background: none; font-weight: bold; } /* operator */
    .pv td.nb { border: none; background: none; }       /* blank, no border */
    .pv tr.sum td.m, .pv tr.sum td.x { border-top: 3px solid #555; } /* sum row separator */
    .pv .cap { font-size: 0.6em; color: #444; text-align: center !important; margin-top: -0.3em; }
    .zt table { border-collapse: collapse; margin: 0.6em auto; font-size: 0.85em; }
    .zt td, .zt th { border: 1px solid #999; padding: 0.3em 0.6em; text-align: center !important; }
    .zt tr.hl td { background: #ffd6d6; color: #b00000; font-weight: bold; }
    </style>
---

# Representations of Numbers in Computer

## Place Value

Every number is a string of digits in boxes; each box has a **place value**.

::: {.pv}
<table>
<tr><th>10³</th><th>10²</th><th>10¹</th><th>10⁰</th></tr>
<tr><td class="m">1</td><td class="m">3</td><td class="m">4</td><td class="m">8</td></tr>
</table>
<p class="cap">$1348 = 1\times10^3 + 3\times10^2 + 4\times10^1 + 8\times10^0$</p>
:::

A computer uses base 2, so each box (a **bit**) holds 0 or 1:

::: {.pv}
<table>
<tr><th>2³</th><th>2²</th><th>2¹</th><th>2⁰</th></tr>
<tr><td class="m">1</td><td class="m">1</td><td class="m">0</td><td class="m">1</td></tr>
</table>
<p class="cap">$1101_2 = 1\times2^3 + 1\times2^2 + 0\times2^1 + 1\times2^0 = 13$</p>
:::

## Integers: Fixed Number of Boxes {.smaller}

An integer is stored in a **fixed** number of bits, e.g. 32 or 64. With $p$ bits the largest unsigned value is
$$
\sum_{i=0}^{p-1} 2^i = 2^p - 1 .
$$

A 4-bit example: the largest storable integer is $1111_2 = 15$.

::: {.pv}
<table>
<tr><th>2⁴</th><th>2³</th><th>2²</th><th>2¹</th><th>2⁰</th></tr>
<tr><td class="x">1</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td></tr>
</table>
<p class="cap">$15 + 1 = 16 = 10000_2$ needs a fifth box that does not exist → **integer overflow**</p>
:::

## Integers: Fixed Number of Boxes (Continued) {.smaller}

R integers use 32 bits with one sign bit, so the largest is $2^{31}-1$:

```{r}
.Machine$integer.max
.Machine$integer.max + 1L    # NA, with a warning
```

Unlike C, R does **not** wrap around silently — it returns `NA`. R also auto-promotes to `double` for most arithmetic, so overflow rarely bites *unless* both operands stay integer:

```{r}
50000L * 50000L   # integer * integer overflows
50000 * 50000      # double: fine
```

## Floating Point Numbers: Scientific Notation in Boxes

A real number is stored as **sign**, **mantissa** (significant digits) and **exponent**:
$$x = (-1)^{s} \times 1.d_1 d_2 \cdots d_p \times 2^{\,e}$$

::: {.pv}
<table>
<tr><th>sign</th><th colspan="4">mantissa (fixed number of digits)</th><th></th><th colspan="2">exponent</th></tr>
<tr><td class="s">+</td><td class="m">1</td><td class="p">.</td><td class="m">d₁</td><td class="m">d₂ … d<sub>p</sub></td><td class="p">×2^</td><td class="s">±</td><td class="e">e</td></tr>
</table>
:::

Double precision (`double` in R) uses 64 bits:

| part | bits | consequence |
|:-|:-:|:----------|
| sign | 1 | |
| mantissa | 52 | about 16 significant decimal digits |
| exponent | 11 | $e \in [-1022,\ 1023]$, i.e. magnitudes from $10^{-308}$ to $10^{308}$ |

## A Toy Decimal Computer: Ruler and Zoom

A floating point number is a **ruler** and a **zoom setting**:

- The mantissa is a ruler with a **fixed number of equally spaced marks**.
- The exponent **stretches** the ruler by $10^{e}$ over $[10^{e}, 10^{e+1})$.
- Stretching adds no marks — the marks **spread apart** as the zoom (exponent) grows.

Our toy computer stores six boxes: sign, a **2-digit mantissa** $d_1.d_2$ ($d_1\in\{1,\ldots,9\}$, $d_2\in\{0,\ldots,9\}$), and a signed 1-digit exponent, $e\in\{-9,\ldots,9\}$:

::: {.pv}
<table>
<tr><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th></th><th>±</th><th>exp</th></tr>
<tr><td class="s">+</td><td class="m">3</td><td class="p">.</td><td class="m">1</td><td class="p">×10^</td><td class="s">+</td><td class="e">0</td></tr>
</table>
<p class="cap">$+3.1\times10^{+0}$ — the closest storable value to $\pi$</p>
:::

## Zoom Levels: How Many Numbers Exist? {.smaller}

Each of the **19 zoom levels** ($e = -9,\ldots,9$) is a ruler of **90 marks** ($1.0, 1.1, \ldots, 9.9$). At the lowest zoom ($e=-9$) the leading digit may also be $0$ — **subnormal** numbers — filling the gap between $0$ and $1.0\times10^{-9}$.

::: {.zt}
<table>
<tr><th>zoom $e$</th><th>interval</th><th>mantissa</th><th>magnitude</th><th>gap</th></tr>
<tr><td>$-9$ (subnormal)</td><td>$(0, 10^{-9})$</td><td>$0.1,\ldots,0.9$</td><td>$\times10^{-9}$</td><td>$10^{-10}$</td></tr>
<tr><td>$-9$</td><td>$[10^{-9}, 10^{-8})$</td><td>$1.0,\ldots,9.9$</td><td>$\times10^{-9}$</td><td>$10^{-10}$</td></tr>
<tr><td>$\vdots$</td><td>$\vdots$</td><td>$\vdots$</td><td>$\vdots$</td><td>$\vdots$</td></tr>
<tr class="hl"><td>$0$</td><td>$[1, 10)$</td><td>$1.0, 1.1, \ldots, 9.9$</td><td>$\times10^{0}$</td><td>$0.1$</td></tr>
<tr><td>$+1$</td><td>$[10, 100)$</td><td>$1.0,\ldots,9.9$</td><td>$\times10^{1}$</td><td>$1$</td></tr>
<tr><td>$\vdots$</td><td>$\vdots$</td><td>$\vdots$</td><td>$\vdots$</td><td>$\vdots$</td></tr>
<tr><td>$+9$</td><td>$[10^{9}, 10^{10})$</td><td>$1.0,\ldots,9.9$</td><td>$\times10^{9}$</td><td>$10^{8}$</td></tr>
</table>
:::


## Visualizing the Zoom

Stretching by $\times10$ multiplies the **gap** by $10$ — it does not add marks:

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.2
par(mfrow = c(2, 1), mar = c(2.5, 1, 1.5, 1))
m0 <- (10:99) / 10; m1 <- 10:99
plot(NA, xlim = c(0, 10.5), ylim = c(0, 1), axes = FALSE, xlab = "", ylab = "",
     main = "e = 0   (gap 0.1)", cex.main = 0.9)
segments(m0, 0.3, m0, 0.7, col = "steelblue"); axis(1, at = 0:10)
plot(NA, xlim = c(0, 105), ylim = c(0, 1), axes = FALSE, xlab = "", ylab = "",
     main = "e = +1   (gap 1)", cex.main = 0.9)
segments(m1, 0.3, m1, 0.7, col = "darkorange3"); axis(1, at = seq(0, 100, 10))
```

The **absolute** gap grows with magnitude; the **relative** gap stays roughly constant across zoom levels.

## Rounding: Relative Error is Constant

A number that falls between two marks is stored as the nearer mark. These three numbers share the same mantissa digits and differ only in exponent (zoom):

| $x$ | stored | error |
|:--|:--:|:--:|
| $0.234$ | $0.23$ | $0.004$ (1.7%) |
| $2.34$ | $2.3$ | $0.04$ (1.7%) |
| $23.4$ | $23$ | $0.4$ (1.7%) |

Absolute error scales with the zoom; **relative** error does not — it is set by how many mantissa digits exist, not by magnitude.

## Rounding in Addition: Machine Epsilon

To add, align by place value. Solid boxes are stored digits, dashed boxes are alignment zeros; digits beyond the mantissa (red) are rounded off.

:::: {.columns}
::: {.column width="33%"}
::: {.pv}
<table>
<tr><th></th><th>10⁰</th><th></th><th>10⁻¹</th><th>10⁻²</th></tr>
<tr><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">0</td><td class="nb"></td></tr>
<tr><td class="op">+</td><td class="z">0</td><td class="p">.</td><td class="m">1</td><td class="m">0</td></tr>
<tr class="sum"><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">1</td><td class="x">0</td></tr>
</table>
<p class="cap">$1.0+0.10=1.1 \ne 1.0$</p>
:::
:::
::: {.column width="33%"}
::: {.pv}
<table>
<tr><th></th><th>10⁰</th><th></th><th>10⁻¹</th><th>10⁻²</th></tr>
<tr><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">0</td><td class="nb"></td></tr>
<tr><td class="op">+</td><td class="z">0</td><td class="p">.</td><td class="m">1</td><td class="m">4</td></tr>
<tr class="sum"><td class="op"></td><td class="m">1</td><td class="p">.</td><td class="m">1</td><td class="x">4</td></tr>
</table>
<p class="cap">$1.0+0.14 \to 1.1$</p>
:::
:::
::: {.column width="33%"}
::: {.pv}
<table>
<tr><th></th><th>10¹</th><th>10⁰</th><th></th><th>10⁻¹</th><th>10⁻²</th></tr>
<tr><td class="op"></td><td class="m">1</td><td class="m">0</td><td class="p">.</td><td class="nb"></td><td class="nb"></td></tr>
<tr><td class="op">+</td><td class="z">0</td><td class="z">0</td><td class="p">.</td><td class="m">4</td><td class="m">0</td></tr>
<tr class="sum"><td class="op"></td><td class="m">1</td><td class="m">0</td><td class="p">.</td><td class="x">4</td><td class="x">0</td></tr>
</table>
<p class="cap">$10+0.40 \to 10$</p>
:::
:::
::::

**Machine epsilon** $\epsilon$ = the gap next to $1$ = the smallest increment the last mantissa box can register: here $\epsilon = 0.1$. Below it, digits are lost — partly (middle) or wholly (right, at a higher zoom).


## Roundoff Error: General Principle {.smaller}

- The stored result of $a+b$ (or $a\times b$, …) is the **nearest representable number** to the exact result — as in the addition examples above.
- Each operation has relative error up to about $\epsilon$; the concern is when these errors **accumulate** or are **amplified**.
- Two patterns amplify roundoff error: adding a **large + small** number, and subtracting **large − large** nearly equal numbers (**catastrophic cancellation**).
- Doubles represent **every integer exactly** up to $2^{53}$; beyond that the gap exceeds 1, so some integers are skipped.

```{r}
#| code-fold: true
.Machine$double.eps      # 2^-52
1 + 1e-16 == 1
2^53 + 1 == 2^53          # gap is now 2: the +1 is lost
```

## Largest and Smallest Toy Numbers

| | sign | mantissa | exponent | value |
|:--|:--:|:--:|:--:|:--:|
| largest | + | $9.9$ | $+9$ | $9.9\times10^{9}$ |
| smallest positive | + | $0.1$ | $-9$ | $0.1\times10^{-9}$ |
| largest negative | − | $0.1$ | $-9$ | $-0.1\times10^{-9}$ |
| smallest | − | $9.9$ | $+9$ | $-9.9\times10^{9}$ |

## Overflow

Adding two large toy numbers can need an exponent the single digit cannot hold:

::: {.pv}
<table>
<tr><th></th><th>10¹⁰</th><th>10⁹</th><th>10⁸</th></tr>
<tr><td class="op"></td><td class="nb"></td><td class="m">6</td><td class="m">0</td></tr>
<tr><td class="op">+</td><td class="nb"></td><td class="m">5</td><td class="m">0</td></tr>
<tr class="sum"><td class="op"></td><td class="x">1</td><td class="m">1</td><td class="x">0</td></tr>
</table>
<p class="cap">$6.0\times10^{9} + 5.0\times10^{9} = 1.1\times10^{+10}$: exponent $10$ needs two digits → **overflow**, stored as `Inf`</p>
:::

For doubles the largest value is $\approx 1.8\times10^{308}$:

```{r}
.Machine$double.xmax
2^1024
```

## Underflow

A tiny result can fall outside the mantissa entirely:

::: {.pv}
<table>
<tr><th></th><th>sign</th><th>1</th><th></th><th>10⁻¹</th><th>10⁻²</th><th></th><th>±</th><th>exp</th></tr>
<tr><td class="nb"></td><td class="s">+</td><td class="m">0</td><td class="p">.</td><td class="x">0</td><td class="x">1</td><td class="p">×10^</td><td class="s">−</td><td class="e">9</td></tr>
</table>
<p class="cap">$(0.1\times10^{-9})\times0.1 = 0.01\times10^{-9}$: the digit $1$ falls outside the mantissa → **underflow**, stored as `0`</p>
:::

For doubles the smallest positive value is $2^{-1074}\approx 4.9\times10^{-324}$ (the smallest with full precision is $2^{-1022}\approx 2.2\times10^{-308}$):

```{r}
.Machine$double.xmin
2^-1075
```


# Rounding Error and Methods for Preventing It

## Computing Sum: Avoiding Large + Small

Compute $1.0 + 0.04 + 0.04$ in our 2-digit-mantissa toy system (exact value $1.08$).

**Left to right**, $(1.0 + 0.04) + 0.04$: each addition needs a third mantissa digit, so each rounds back to $1.0$ (the same loss as $1.0+0.40\to1.0$ shown earlier, just smaller). Both small terms are lost — result $1.0$.

**Small terms first**, $1.0 + (0.04 + 0.04)$:

::: {.pv}
<table>
<tr><td class="m">8</td><td class="p">.</td><td class="m">0</td><td class="p">×10^</td><td class="s">−</td><td class="e">2</td>
<td class="p" style="padding:0 0.6em">then 1.0 + 0.08 →</td>
<td class="m">1</td><td class="p">.</td><td class="m">1</td></tr>
</table>
<p class="cap">$0.04+0.04=0.08$ exactly, then $1.0+0.08\to1.1$ — one roundoff, not an accumulating one.</p>
:::

## Computing Sum: Why Order Matters

Summing $n$ terms in a loop `s <- s + x[i]`: once `s` is large, every small `x[i]` is partly or wholly rounded away, and the errors accumulate over $n$ additions.

**Rules for accurate summation**

- Add numbers of **similar magnitude** together; add the smallest terms first (sort ascending).
- **Pairwise summation**: sum halves recursively, so partial sums stay of similar size (used by R's `sum()`).
- **Kahan compensated summation**: carry the rounded-away part to the next step.

```{r}
x <- c(1, rep(1e-16, 1e4))
s <- 0; for (v in x) s <- s + v      # naive loop
c(loop = s - 1, sorted = sum(sort(x)) - 1, exact = 1e-12)
```

## Kahan Compensated Summation

Track the part rounded away at each step, and add it back at the next step:

```{r}
kahan_sum <- function(x) {
  s <- 0; c <- 0                       # c holds the running roundoff error
  for (v in x) { y <- v - c; t <- s + y; c <- (t - s) - y; s <- t }
  s
}
c(kahan = kahan_sum(x) - 1, builtin = sum(x) - 1, exact = 1e-12)
```

Kahan summation and R's `sum()` (pairwise + extended precision) both recover the answer to full precision, without needing to sort the data first.

## Computing $e^x$: Avoiding Large − Large

Taylor series:
$$e^x \approx \sum_{k=0}^{K}\frac{x^k}{k!} = 1 + x + \frac{x^2}{2!} + \frac{x^3}{3!} + \cdots$$

For $x = -20$ the terms alternate in sign and are huge relative to the answer $e^{-20}\approx 2\times10^{-9}$:

| $k$ | 0 | 1 | 2 | … | 20 | … |
|:--|:--:|:--:|:--:|:--:|:--:|:--:|
| $x^k/k!$ | $1$ | $-20$ | $200$ | … | $4.3\times10^{7}$ | … |

Each term carries roundoff of relative size $\epsilon \approx 10^{-16}$, i.e. absolute size $\approx 4\times10^{-9}$ for the largest terms — **as large as the answer itself**. Subtracting nearly equal large numbers leaves only their errors: **catastrophic cancellation**.

## Computing $e^x$: Fix by Reformulating

For $x>0$ all terms are positive, so no cancellation occurs. Therefore compute
$$e^{-x} = \frac{1}{e^{x}}, \qquad x > 0 .$$

```{r}
taylor_exp <- function(x, K = 100) sum(x^(0:K) / factorial(0:K))
c(naive = taylor_exp(-20), reciprocal = 1 / taylor_exp(20), true = exp(-20))
```

**General principle**: rearrange a formula so that quantities of similar magnitude and **the same sign** are combined; avoid computing a small result as the difference of two large numbers.

## Computing Variance: Avoiding Large − Large

Two algebraically identical formulas for the sum of squared deviations:
$$\underbrace{\sum_{i=1}^{n}(x_i - \bar{x})^2}_{\text{two-pass}}
\;=\;
\underbrace{\sum_{i=1}^{n} x_i^2 - \frac{\left(\sum_i x_i\right)^2}{n}}_{\text{one-pass}}$$

- The one-pass formula needs only $\sum x_i$ and $\sum x_i^2$, computed in a single sweep of the data.
- When the $x_i$ are large but close together, $\sum x_i^2$ and $(\sum x_i)^2/n$ are both huge and nearly equal: **large − large**.

## Computing Variance: Cancellation Illustrated

$x = (1, 2, 3)$: deviations $(-1, 0, 1)$, so $\sum(x_i-\bar x)^2 = 2$.

Shift by $M = 10^{8}$: $x = (M+1, M+2, M+3)$. The deviations and the answer are **unchanged**, but the one-pass formula computes

$$\sum x_i^2 \approx 3\times10^{16}, \qquad \frac{(\sum x_i)^2}{n} \approx 3\times10^{16}, \qquad \text{difference} = 2.$$

With 16 significant digits, $3\times10^{16}$ is stored with a gap of about $4$ between neighbours:

::: {.pv}
<table>
<tr><th colspan="16">16 mantissa digits</th><th>lost</th></tr>
<tr><td class="m">3</td><td class="p">.</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">1</td><td class="m">2</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">0</td><td class="m">1</td><td class="x">4</td><td class="p">×10^16</td></tr>
</table>
<p class="cap">The true answer, $2$, lives entirely in the lost digits.</p>
:::

```{r}
x <- c(1, 2, 3) + 1e8
c(one_pass = sum(x^2) - sum(x)^2 / 3, two_pass = sum((x - mean(x))^2))
```

## Computing Variance: Gradual Failure

The one-pass error grows like $\epsilon M^2 n$, so it worsens continuously as the shift $M$ grows — eventually giving a **negative** "variance", which is impossible:

```{r}
var1 <- function(x) { n <- length(x); (sum(x^2) - sum(x)^2 / n) / (n - 1) }
M <- 10^(0:10)
t(sapply(M, function(m) { x <- c(1, 2, 3) + m; c(M = m, one_pass = var1(x)) }))
```

## Computing Variance: Fixes

- **Two-pass**: compute $\bar x$ first, then $\sum (x_i - \bar x)^2$. Deviations are small and of similar size, so no cancellation. (`var()` does this.)
- **Shifted one-pass**: subtract any convenient constant $c \approx \bar x$ (e.g. $c = x_1$) before applying the one-pass formula:
$$\sum (x_i-\bar x)^2 = \sum (x_i - c)^2 - \frac{\left(\sum (x_i - c)\right)^2}{n}$$
- **Welford's online algorithm** updates $\bar x_k$ and $\sum_{i\le k}(x_i - \bar x_k)^2$ one observation at a time; use it for streaming data.

```{r}
var_shifted <- function(x) { n <- length(x); d <- x - x[1]; (sum(d^2) - sum(d)^2 / n) / (n - 1) }
var_welford <- function(x) {
  m <- 0; S <- 0
  for (k in seq_along(x)) { m_old <- m; m <- m + (x[k] - m) / k; S <- S + (x[k] - m_old) * (x[k] - m) }
  S / (length(x) - 1)
}
x <- c(1, 2, 3) + 1e10
c(one_pass = var1(x), shifted = var_shifted(x), welford = var_welford(x))
```

# Methods for Mitigating Overflow and Underflow Errors

## Overflow and Underflow in Statistical Computing

Likelihoods are **products of many probabilities**; normalizing constants involve **factorials** or **exponentials**.

$$\prod_{i=1}^{2000} p(x_i) = \left(\tfrac12\right)^{2000} \approx 10^{-602} \quad\to\quad \text{underflow to } 0$$

$$1000! \approx 4\times10^{2567} \quad\to\quad \text{overflow to } \infty$$

$$\text{logistic: } \frac{e^{x}}{1+e^{x}} \text{ at } x = 1000 \quad\to\quad \frac{\infty}{\infty} = \texttt{NaN}$$

Two remedies: (1) **change the algebraic expression**; (2) **work on the log scale**.

## Changing Algebraic Expression: The Sigmoid Function

Two equivalent forms:
$$\sigma(x) = \frac{e^{x}}{1+e^{x}} = \frac{1}{1+e^{-x}}$$

| $x$ | $e^{x}/(1+e^{x})$ | $1/(1+e^{-x})$ |
|:--:|:--|:--|
| $+1000$ | $e^{x}=\infty$, $\infty/\infty = $ `NaN` ✗ | $e^{-x}=0$, $1/1 = 1$ ✓ |
| $-1000$ | $e^{x}=0$, $0/1 = 0$ ✓ | $e^{-x}=\infty$, $1/\infty = 0$ ✓ |

The second form never divides $\infty$ by $\infty$. Equivalently, pick the form where the exponential has a **non-positive** argument:
$$\sigma(x) = \begin{cases} 1/(1+e^{-x}) & x \ge 0\\ e^{x}/(1+e^{x}) & x < 0\end{cases}$$

```{r}
x <- 1000
c(naive = exp(x) / (1 + exp(x)), stable = 1 / (1 + exp(-x)))
```

## Changing Algebraic Expression: Built-in Functions {.smaller}

Software provides functions written specifically to avoid cancellation and overflow.

| instead of | use | reason |
|:--|:--|:--|
| `log(1 + x)` | `log1p(x)` | for tiny $x$, `1 + x` rounds to 1 |
| `exp(x) - 1` | `expm1(x)` | for tiny $x$, `exp(x)` rounds to 1 |
| `log(factorial(n))` | `lfactorial(n)`, `lgamma(n + 1)` | `factorial(n)` overflows |
| `log(choose(n, k))` | `lchoose(n, k)` | overflow |
| `log(dnorm(x))` | `dnorm(x, log = TRUE)` | `dnorm(x)` underflows |
| `log(pnorm(x))` | `pnorm(x, log.p = TRUE)` | underflow |
| `log(1 - pnorm(x))` | `pnorm(x, lower.tail = FALSE, log.p = TRUE)` | cancellation *and* underflow |

```{r}
#| code-fold: true
c(log(1 + 1e-20), log1p(1e-20))
c(log(factorial(1000)), lfactorial(1000))
c(naive = log(1 - pnorm(10)), stable = pnorm(10, lower.tail = FALSE, log.p = TRUE))
```

`1 - pnorm(10)` rounds to exactly `1 - 1 = 0` in double precision, so the naive answer is `-Inf`; the true value is about $-53.2$.

## Representing Numbers in Logarithmic Form

Represent a positive number $x$ by $\ell_x = \log x$:
$$x = e^{2000} \iff \ell_x = 2000, \qquad x = e^{-2000} \iff \ell_x = -2000 .$$

Numbers that overflow or underflow in ordinary form are ordinary numbers on the log scale.

We then need each operation $f$ **translated** to the log scale:
$$x_1,\dots,x_n \xrightarrow{\;f\;} y \qquad\Longleftrightarrow\qquad \ell_1,\dots,\ell_n \xrightarrow{\;\tilde f\;} \ell_y = \log f(e^{\ell_1},\dots,e^{\ell_n})$$

## Products and Quotients on the Log Scale

$$\log(x_1 x_2) = \ell_1 + \ell_2, \qquad \log\left(\prod_{i=1}^{n} x_i\right) = \sum_{i=1}^{n} \ell_i, \qquad \log\frac{x_1}{x_2} = \ell_1 - \ell_2$$

**Log-likelihood**: $\log \prod_i p(x_i) = \sum_i \log p(x_i)$ — always compute the log-likelihood, never the likelihood.

**Example**: $x_1 = 2^{2000}$, $x_2 = 2^{2001}$ both overflow, but
$$\log\frac{x_1}{x_2} = 2000\log 2 - 2001\log 2 = -\log 2 .$$

```{r}
c(direct = 2^2000 / 2^2001, log_scale = exp(2000 * log(2) - 2001 * log(2)))
```

## The Log-Sum-Exp Trick

Sums are the hard case: $\log(x_1 + x_2) \neq \ell_1 + \ell_2$. We need
$$\log\left(\sum_{i=1}^{n} x_i\right) = \log\left(\sum_{i=1}^{n} e^{\ell_i}\right),$$
but the $e^{\ell_i}$ may overflow or underflow. Let $m = \max_i \ell_i$ and factor it out:
$$\log\left(\sum_{i=1}^{n} e^{\ell_i}\right)
= \log\left(e^{m}\sum_{i=1}^{n} e^{\ell_i - m}\right)
= m + \log\left(\sum_{i=1}^{n} e^{\ell_i - m}\right)$$

- Every $\ell_i - m \le 0$, so $e^{\ell_i - m} \in (0, 1]$: **no overflow**.
- The largest term is $e^{0} = 1$, so the sum is at least 1: **no underflow** of the total.

## Log-Sum-Exp: Illustration

$x_i = e^{2000}, e^{2001}, e^{2002}$ — each overflows to `Inf`, so $\sum x_i$ computed directly is `Inf`.

| | $i=1$ | $i=2$ | $i=3$ |
|:--|:--:|:--:|:--:|
| $\ell_i$ | 2000 | 2001 | 2002 |
| $\ell_i - m$ ($m = 2002$) | $-2$ | $-1$ | $0$ |
| $e^{\ell_i - m}$ | $0.135$ | $0.368$ | $1$ |

$$\log\sum_i x_i = 2002 + \log(0.135 + 0.368 + 1) = 2002.408$$

```{r}
log_sum_exp <- function(l) { m <- max(l); m + log(sum(exp(l - m))) }
c(direct = log(sum(exp(c(2000, 2001, 2002)))), lse = log_sum_exp(c(2000, 2001, 2002)))
```

Applications: marginal likelihoods $\log\sum_k \pi_k p_k(x)$ in mixtures, posterior normalizing constants, softmax.

## Difference on the Log Scale

For $x > y > 0$ with $\ell_x = \log x$, $\ell_y = \log y$:
$$\log(x - y) = \log\left(e^{\ell_x}\left(1 - e^{\ell_y - \ell_x}\right)\right) = \ell_x + \log\left(1 - e^{\ell_y - \ell_x}\right)$$

- Since $\ell_y - \ell_x < 0$, $e^{\ell_y-\ell_x}\in(0,1)$: no overflow.
- When $x \approx y$, use `log1p(-exp(ly - lx))` to avoid the cancellation in $1 - e^{\ell_y-\ell_x}$.

## Summary

- Integers occupy a fixed number of bits; exceeding $2^{31}-1$ gives `NA` in R (not silent wraparound as in C).
- Doubles store sign + 52-bit mantissa (~16 decimal digits) + 11-bit exponent ($10^{-308}$ to $10^{308}$); the relative spacing everywhere is $\epsilon = 2^{-52}$, and integers above $2^{53}$ are not all representable.
- **Overflow** → `Inf`; **underflow** → `0` (silent, and thus more dangerous).
- Roundoff is amplified by **large + small** (fix: add small terms first, pairwise/Kahan summation) and **large − large** (fix: reformulate to combine same-sign, similar-magnitude quantities — $e^{-x}=1/e^x$, two-pass variance).
- Overflow/underflow are avoided by choosing non-positive exponent arguments, using purpose-built functions (`log1p`, `expm1`, `lgamma`, `dnorm(log=TRUE)`), and working on the **log scale** with the log-sum-exp trick.

## Practice Questions

**Question 1.** How do you compute
$$\log\left(\frac{1}{1+e^{-x}}\right)$$
so as to avoid overflow and underflow of $e^{-x}$, for both large positive and large negative $x$?

**Question 2 (log-softmax).** How do you compute
$$\log\left(\frac{e^{x_k}}{\sum_{j=1}^{K} e^{x_j}}\right)$$
so as to avoid underflow and overflow of $e^{x_j}$?

**Question 3.** The one-pass variance formula fails for $x = (M+1, M+2, M+3)$ with $M = 10^{8}$. Predict the smallest $M$ (a power of 10) at which the one-pass result in R first differs from $2$, then verify.

## Practice Questions (cont.)

**Question 4.** Write `log_dbinom(k, n, p)` returning $\log\binom{n}{k}p^k(1-p)^{n-k}$ for $n = 10^5$ without overflow or underflow, using `lchoose` and `log1p`. Compare with `dbinom(k, n, p, log = TRUE)`.

**Question 5.** Implement Welford's algorithm for the sample **covariance** of two variables, and check it against `cov()` on data shifted by $10^{10}$.
