---
subtitle: "Importance Sampling"
author: "Longhai Li"
format: banner_slides-revealjs
number-sections: false
---

```{r}
#| include: false
set.seed(812)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
library(latex2exp)
ess <- function(w) sum(w)^2 / sum(w^2)
```

# 1. Goal and Idea

## The Goal

Two integrals with respect to an **unnormalized** function $f(\theta)$, e.g. $f(\theta) = \pi(\theta)P(x\mid\theta)$:

1. A posterior expectation
$$A = \frac{\int a(\theta)\,f(\theta)\,d\theta}{\int f(\theta)\,d\theta} = E_f\big[a(\theta)\big], \qquad \theta\sim\frac{f(\theta)}{C_f}$$

2. The normalizing constant (marginal likelihood)
$$C_f = \int f(\theta)\,d\theta$$

Plain Monte Carlo needs draws from $f/C_f$, which we cannot produce. Rejection sampling needs an envelope $g\ge f$, which may be hard to find or very wasteful.

## The Idea: Reweighting

*Reject, reweight, repeat.* Instead of rejecting proposals, keep all of them and **weight** each by how much $f$ favours it relative to the proposal.

- Draw $\theta\sim g(\theta)/C_g$, where $g$ is another (possibly unnormalized) function we can sample from
- Attach the **importance weight**
$$w(\theta) = \frac{f(\theta)}{g(\theta)}$$
- Hope that $w(\theta)\approx$ constant, i.e. $g$ has roughly the shape of $f$

**Requirement on $g$**: wherever $f$ is positive, $g$ must be positive too,
$$\{\theta: g(\theta)>0\}\ \supseteq\ \{\theta: f(\theta)>0\}$$

Otherwise part of $f$ is never visited and the estimate is biased (Example 3).

# 2. The Importance Reweighting Formula

## Derivation

Let $C_g = \int g(\theta)\,d\theta$. For any $a$,
$$
\begin{aligned}
E_f\big[a(\theta)\big] &= \int a(\theta)\,\frac{f(\theta)}{C_f}\,d\theta
= \frac{C_g}{C_f}\int a(\theta)\,\frac{f(\theta)}{g(\theta)}\cdot\frac{g(\theta)}{C_g}\,d\theta \\[4pt]
&= \frac{C_g}{C_f}\,E_g\big[a(\theta)\,w(\theta)\big], \qquad w(\theta) = \frac{f(\theta)}{g(\theta)}
\end{aligned}
$$

The division by $g$ is legitimate only on $\{g>0\}$, hence the support requirement.

Taking $a\equiv1$,
$$\frac{C_f}{C_g} = \int\frac{f(\theta)}{g(\theta)}\cdot\frac{g(\theta)}{C_g}\,d\theta = E_g\big[w(\theta)\big]
= \frac{\int f(\theta)\,d\theta}{\int g(\theta)\,d\theta}$$

Combining,
$$E_f\big[a(\theta)\big] = \frac{E_g\big[a(\theta)\,w(\theta)\big]}{E_g\big[w(\theta)\big]}$$

## The Estimators

Draw $\theta_1,\ldots,\theta_n \overset{iid}{\sim} g/C_g$ and compute $w_i = w(\theta_i) = f(\theta_i)/g(\theta_i)$.

**Normalizing constant ratio**
$$\widehat{\frac{C_f}{C_g}} = \frac1n\sum_{i=1}^n w_i$$

If $g$ is a normalized density ($C_g = 1$), this estimates $C_f$ itself, e.g. the marginal likelihood.

**Expectation** (self-normalized importance sampling)
$$\hat A = \frac{\frac1n\sum_{i=1}^n a(\theta_i)\,w_i}{\frac1n\sum_{i=1}^n w_i}
= \sum_{i=1}^n a(\theta_i)\,\tilde w_i, \qquad
\tilde w_i = \frac{w_i}{\sum_{j=1}^n w_j}, \quad \sum_{i=1}^n\tilde w_i = 1$$

The normalized weights $\tilde w_i$ turn the proposal sample into a **weighted sample from $f$**; unknown constants in $f$ and $g$ cancel.

Example: $w = (100, 40, 60)$ gives $\tilde w = (0.5, 0.2, 0.3)$.

## Interpretation and Properties

- $\hat A$ is a weighted average of $a(\theta_i)$: draws in regions where $f\gg g$ count more, draws where $f\ll g$ count less
- $\frac1n\sum_i w_i$ is unbiased for $C_f/C_g$; $\hat A$ is a ratio of two averages, so it is biased with bias $O(1/n)$ but consistent
- Nothing is discarded, unlike rejection sampling; no envelope condition $g\ge f$ is needed
- Everything hinges on the variance of the weights: if $w(\theta)$ varies wildly, a few draws dominate the sum

```{r}
#| fig-width: 9
#| fig-height: 3.4
par(mar = c(4, 4, 1, 1))
f <- function(t) dgamma(t, 4, 1.5); g <- function(t) dnorm(t, 2.5, 1.6)
th <- rnorm(30, 2.5, 1.6); w <- f(th) / g(th); wt <- w / sum(w)
curve(f, -1, 8, lwd = 3, col = "firebrick", ylab = "", xlab = TeX(r"($\theta$)"), ylim = c(0, 0.45))
curve(g, add = TRUE, lwd = 3, col = "steelblue")
points(th, rep(0, 30), pch = 19, col = adjustcolor("gray20", 0.6), cex = 0.6 + 12 * wt)
legend("topright", bty = "n", lwd = 3, col = c("firebrick", "steelblue"), legend = TeX(c(r"(target $f$)", r"(proposal $g$)")))
mtext(TeX(r"(dot size $\propto$ normalized weight $\tilde{w}_i$)"), side = 3, line = -1.2, adj = 0.02, cex = 0.9)
```

## Effective Sample Size

How many "equally informative" draws is a weighted sample worth?
$$S_{\text{eff}} = \frac{1}{\sum_{s=1}^S \tilde w(\theta_s)^2}, \qquad \tilde w(\theta_s) = \frac{w(\theta_s)}{\sum_{j=1}^S w(\theta_j)}$$

**Special case**: if $w(\theta_s)\approx c$ for all $s$ (proposal has the shape of the target), then $\tilde w(\theta_s) = 1/S$ and
$$S_{\text{eff}} = \frac{1}{\sum_{s=1}^S (1/S)^2} = \frac{1}{1/S} = S$$

At the other extreme, one dominant weight gives $S_{\text{eff}}\approx1$.

**Variance heuristic.** For iid draws from $f$, $\text{Var}_f\big(\tfrac1S\sum_s a(\theta_s)\big) = \text{Var}_f(a)/S$. For the importance estimator,
$$\text{Var}_g\left(\frac{\sum_s a(\theta_s)w(\theta_s)}{\sum_s w(\theta_s)}\right)\ \approx\ \frac{\text{Var}_f\big(a(\theta)\big)}{S_{\text{eff}}}$$

We will see (Examples 2 and 3) that $S_{\text{eff}}$ is computed only from the weights **observed**, and can look excellent when the estimate is wrong.

# 3. A Good Story

## Example 1: Estimating a Normal Tail Probability

Estimate
$$I = \int_a^b \varphi(x)\,dx = P(a<X<b), \qquad X\sim N(0,1)$$

and the conditional expectation $E\big[a(X)\mid a<X<b\big]$. Set the target to the **truncated** density
$$f(x) = \varphi(x)\,I(a\le x\le b), \qquad C_f = \int_{-\infty}^{\infty}\varphi(x)\,I(a\le x\le b)\,dx = P(a<X<b)$$

so that
$$E_f\big[a(x)\big] = \int_a^b a(x)\,\frac{\varphi(x)}{I}\,dx = E\big[a(X)\mid a<X<b\big]$$

For a far tail ($a = 4$), plain Monte Carlo from $N(0,1)$ almost never lands in $[a,b]$ (probability $\approx 3\times10^{-5}$), so both $C_f$ and $E_f[a]$ are hopeless without a proposal concentrated on $[a,b]$.

## Example 1 (Continued): Choice of Proposal

Any normalized $g$ supported on $[a,b]$ works, and then $\frac1n\sum_i w_i$ estimates $I$ directly:

- **Uniform** $g(x) = \dfrac{I(a\le x\le b)}{b-a}$: $\;w(x) = (b-a)\,\varphi(x)$, so $\hat I = (b-a)\,\frac1n\sum_i\varphi(x_i)$, the Monte Carlo version of the midpoint rule
- **Shifted exponential** $g(x) = \dfrac{\lambda e^{-\lambda(x-a)}}{1-e^{-\lambda(b-a)}}\,I(a\le x\le b)$ with $\lambda\approx a$: matches the decay of $\varphi$ on the tail, so $w$ is nearly constant and $S_{\text{eff}}\approx n$

```{r}
#| echo: true
a <- 4; b <- 6; n <- 1e4; exact <- pnorm(b) - pnorm(a)
x_u <- runif(n, a, b);  w_u <- (b - a) * dnorm(x_u)
lam <- a; x_e <- a - log(1 - runif(n) * (1 - exp(-lam * (b - a)))) / lam       # inverse CDF
g_e <- lam * exp(-lam * (x_e - a)) / (1 - exp(-lam * (b - a))); w_e <- dnorm(x_e) / g_e
rbind(uniform = c(I_hat = mean(w_u), rel_error = mean(w_u) / exact - 1, S_eff = ess(w_u),
                  E_X_given = sum(x_u * w_u) / sum(w_u)),
      exponential = c(mean(w_e), mean(w_e) / exact - 1, ess(w_e), sum(x_e * w_e) / sum(w_e)))
```

## Example 1 (Continued)

```{r}
#| fig-width: 10
#| fig-height: 4.2
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
curve(dnorm, -4, 7, lwd = 3, xlab = "x", ylab = "", main = "target: truncated normal on [a, b]")
xs <- seq(a, b, length.out = 100)
polygon(c(xs, rev(xs)), c(dnorm(xs), rep(0, 100)), col = adjustcolor("firebrick", 0.5), border = NA)
abline(v = c(a, b), lty = 3); text(a, 0.2, TeX(r"(area $= I \approx 3 \times 10^{-5}$)"), pos = 4, col = "firebrick")
g_e_fun <- function(x) lam * exp(-lam * (x - a)) / (1 - exp(-lam * (b - a)))
curve(dnorm(x) / exact, a, b, lwd = 3, col = "firebrick", xlab = "x", ylab = "density", main = "normalized target vs proposals on [a, b]", ylim = c(0, 4.5))
curve(g_e_fun, add = TRUE, lwd = 3, col = "forestgreen"); segments(a, 1 / (b - a), b, 1 / (b - a), lwd = 3, col = "steelblue")
legend("topright", bty = "n", lwd = 3, col = c("firebrick", "forestgreen", "steelblue"),
       legend = TeX(c(r"($f/C_f$)", r"(shifted exponential $g$)", r"(uniform $g$)")))
```

The exponential proposal tracks the shape of the truncated normal; the uniform proposal wastes draws near $b$ where $\varphi$ is tiny, and its weights vary by a factor $e^{(b^2-a^2)/2}\approx e^{10}$.

# 4. Two Bad Stories

## Example 2: Same Support, but a Mode Is Missed

$$f(\theta) = \tfrac12\varphi(\theta+5) + \tfrac12\varphi(\theta-5), \qquad g(\theta) = \varphi(\theta+5)$$

Both are positive on all of $\mathbb R$, so the support requirement holds. But $g$ concentrates around $-5$ and never produces draws near the second mode at $+5$.

For $\theta$ near $-5$,
$$w(\theta) = \frac{f(\theta)}{g(\theta)} = \tfrac12 + \tfrac12\,\frac{\varphi(\theta-5)}{\varphi(\theta+5)} = \tfrac12 + \tfrac12 e^{10\theta}\ \approx\ \tfrac12$$

so all observed weights are nearly equal: $\tilde w_i\approx 1/S$ and $S_{\text{eff}}\approx S$. Yet

- $\widehat{C_f/C_g} = \frac1S\sum_i w_i \approx \tfrac12$, while the truth is $1$
- $\hat E_f(\theta) \approx -5$, while the truth is $0$

The draws that would reveal the error (those with $w\approx \tfrac12 e^{10\theta}\gg1$, near $\theta = 5$) have probability $\approx \Phi(-10)\approx 10^{-23}$ under $g$: they never appear.

## Example 2 (Continued)

```{r}
#| fig-width: 10
#| fig-height: 4.2
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1)); S <- 5000
f2 <- function(t) 0.5 * dnorm(t + 5) + 0.5 * dnorm(t - 5); g2 <- function(t) dnorm(t + 5)
th <- rnorm(S, -5); w <- f2(th) / g2(th); wt <- w / sum(w)
curve(f2, -9, 9, lwd = 3, col = "firebrick", xlab = TeX(r"($\theta$)"), ylab = "", main = "target f and proposal g", ylim = c(0, 0.42))
curve(g2, add = TRUE, lwd = 3, col = "steelblue"); rug(th[1:300], col = adjustcolor("steelblue", 0.5))
legend("topright", bty = "n", lwd = 3, col = c("firebrick", "steelblue"), legend = c("f", "g"))
hist(wt * S, breaks = 40, col = "gray85", border = "white", xlab = TeX(r"($S \tilde{w}_i$)"),
     main = sprintf("weights: S_eff = %.0f of %d", ess(w), S))
mtext(sprintf("C_f/C_g estimate = %.3f (true 1);  E(theta) estimate = %.2f (true 0)", mean(w), sum(th * wt)),
      side = 1, line = 4, adj = 1, cex = 0.85, outer = TRUE)
```

**Lesson**: near-constant weights show only that $g$ matches $f$ **where $g$ has mass**. They say nothing about regions $g$ never visits. Use a proposal wide enough to cover all modes, or a mixture proposal, and compare results across different $g$.

## Example 3: Violating the Support Requirement

$$f(\theta) = I(0\le\theta\le2), \qquad g(\theta) = I(0\le\theta\le1)$$

Here $\text{supp}(g)\subset\text{supp}(f)$: the proposal never visits $(1,2]$. True ratio: $C_f/C_g = 2/1 = 2$.

Draw $\theta_s\sim g$, i.e. $\text{Unif}(0,1)$. For every draw, $w(\theta_s) = f(\theta_s)/g(\theta_s) = 1$, so
$$\widehat{\frac{C_f}{C_g}} = \frac1S\sum_{s=1}^S w(\theta_s) = 1 \ \ne\ 2, \qquad S_{\text{eff}} = S$$

The estimate is wrong **regardless of $S$**, with perfect-looking weights. Example 2 is the smooth version of the same failure. The reverse situation ($g$ wider than $f$) is harmless: draws outside $\text{supp}(f)$ simply receive weight $0$.

```{r}
#| fig-width: 8
#| fig-height: 2.4
par(mar = c(4, 1, 1, 1))
plot(NA, xlim = c(-0.2, 2.4), ylim = c(0, 1.5), xlab = TeX(r"($\theta$)"), ylab = "", yaxt = "n")
rect(0, 0, 2, 1, col = adjustcolor("firebrick", 0.25), border = "firebrick", lwd = 3)
rect(0, 0, 1, 1, col = adjustcolor("steelblue", 0.35), border = "steelblue", lwd = 3)
text(1.5, 0.5, "never sampled", col = "firebrick"); text(0.5, 1.2, TeX(r"($g$: $w = 1$ everywhere)"), col = "steelblue")
text(2, 1.2, TeX(r"($f = I(0 \leq \theta \leq 2)$)"), col = "firebrick")
```

## The Opposite Failure: Heavy-Tailed Weights

Examples 2 and 3 fail silently because $g$ is too **narrow**. A $g$ with lighter tails than $f$ fails noisily: $w = f/g$ is unbounded, $\text{Var}_g(w)$ may be infinite, and a rare tail draw arrives with a huge weight.

```{r}
#| fig-width: 10
#| fig-height: 3.6
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1)); S <- 2000
f3 <- function(t) dt(t, 3)
for (case in 1:2) {
  if (case == 1) { th <- rnorm(S, 0, 1.2); g3 <- function(t) dnorm(t, 0, 1.2); lab <- "proposal N(0, 1.2^2)" }
  else           { th <- 1.5 * rt(S, 3);    g3 <- function(t) dt(t / 1.5, 3) / 1.5; lab <- "proposal 1.5 x t(3)" }
  w <- f3(th) / g3(th); wt <- w / sum(w)
  plot(sort(wt, decreasing = TRUE)[1:200], type = "h", lwd = 2, col = c("firebrick", "steelblue")[case],
       xlab = "draw (sorted by weight)", ylab = TeX(r"($\tilde{w}_i$)"),
       main = sprintf("%s:  S_eff = %.0f, max weight = %.3f", lab, ess(w), max(wt)))
}
```

**Practical rules**: give $g$ tails at least as heavy as $f$ (a $t$ rather than a normal); centre and inflate $g$ using a mode and Hessian (a Laplace approximation is a natural start); report $S_{\text{eff}}$ and $\max_i\tilde w_i$; rerun with different $g$ and distrust diverging answers.

# 5. Two Advanced Applications

## Example 4: Free Energy and the Marginal Likelihood {.smaller}

In statistical physics, a system with energy $U(x)$ at inverse temperature $\beta$ has
$$Z(\beta) = \int e^{-\beta U(x)}\,dx, \qquad F(\beta) = -\frac1\beta\log Z(\beta) \ \ (\text{free energy})$$

Importance sampling from $\beta_0$ to $\beta_1$ with $g = e^{-\beta_0 U}$, $f = e^{-\beta_1 U}$, $w(x) = e^{-(\beta_1-\beta_0)U(x)}$:
$$\frac{Z(\beta_1)}{Z(\beta_0)} = E_{\beta_0}\Big[e^{-(\beta_1-\beta_0)U(x)}\Big]$$

**Bayesian analogue.** With $U(\theta) = -\log L(\theta)$ and $g = \pi(\theta)$ (prior, $\beta_0 = 0$), $f = L(\theta)\pi(\theta)$ (posterior, $\beta_1 = 1$):
$$P(y) = \frac{C_f}{C_g} = E_\pi\big[L(\theta)\big] \approx \frac1S\sum_{s=1}^S L(\theta_s), \qquad \theta_s\sim\pi$$

$-\log P(y)$ is the free energy of the posterior. This "prior sampling" estimator is exactly Example 2 in disguise: the prior is diffuse, the likelihood is a narrow spike, weights are $L(\theta_s)$, and almost all prior draws have negligible weight.

## Example 4 (Continued): Prior Sampling Fails, Tempering Helps

```{r}
#| echo: true
y <- rnorm(30, 1, 1); n <- length(y)                       # data; model N(theta, 1), prior N(0, 10^2)
logL <- function(th) sapply(th, function(t) sum(dnorm(y, t, log = TRUE)))
s2p <- 100; ybar <- mean(y)                                # exact log P(y): y ~ N(0, I + s2p * 11')
exact <- -n / 2 * log(2 * pi) - 0.5 * sum((y - ybar)^2) - 0.5 * log(1 + n * s2p) - 0.5 * n * ybar^2 / (1 + n * s2p)
S <- 1e5; lse <- function(l) { M <- max(l); M + log(sum(exp(l - M))) }
th <- rnorm(S, 0, 10); lw <- logL(th)                       # prior sampling: log weights = log L
c(exact = exact, prior_sampling = lse(lw) - log(S), S_eff = ess(exp(lw - max(lw))))
```

```{r}
#| echo: true
betas <- seq(0, 1, by = 0.1); logZ <- 0                     # stepping stones: chain of small steps
for (k in 2:length(betas)) {                                # draw from tempered posterior beta_{k-1} (conjugate here)
  b0 <- betas[k - 1]; v <- 1 / (1 / s2p + b0 * n); m <- v * b0 * n * ybar
  th <- rnorm(S, m, sqrt(v)); lw <- (betas[k] - b0) * logL(th)
  logZ <- logZ + lse(lw) - log(S) }
c(exact = exact, stepping_stones = logZ)
```

Each small temperature step has near-constant weights (a good story), and the log ratios add up: $\log Z(1) - \log Z(0) = \sum_k \log\frac{Z(\beta_k)}{Z(\beta_{k-1})}$.

## Example 5: Bayesian Cross-Validation

Leave-one-out predictive density, the Bayesian analogue of cross-validation error:
$$p(y_i\mid y_{-i}) = \int p(y_i\mid\theta)\,p(\theta\mid y_{-i})\,d\theta$$

Refitting the model $n$ times is expensive. Instead, reuse draws $\theta_s\sim p(\theta\mid y)$ from the **full** posterior as the proposal. The target is the leave-one-out posterior,
$$f(\theta) = p(\theta\mid y_{-i}) \propto \frac{p(\theta\mid y)}{p(y_i\mid\theta)}, \qquad
w_i(\theta) = \frac{f(\theta)}{g(\theta)} = \frac{1}{p(y_i\mid\theta)}$$

Then
$$\hat p(y_i\mid y_{-i}) = \frac{\sum_s p(y_i\mid\theta_s)\,w_i(\theta_s)}{\sum_s w_i(\theta_s)}
= \frac{S}{\sum_{s=1}^S 1/p(y_i\mid\theta_s)}$$

the harmonic mean of $p(y_i\mid\theta_s)$ over posterior draws (Gelfand, Dey & Chang, 1992).

## Example 5 (Continued): When It Works and When It Breaks

- The full posterior is **narrower** than the leave-one-out posterior (it has one more observation), so $g$ has lighter tails than $f$: the heavy-tailed-weight failure
- For a typical $y_i$ the two posteriors nearly coincide and the weights are stable
- For an **influential** $y_i$ (an outlier, a high-leverage point) the weights $1/p(y_i\mid\theta_s)$ have infinite variance and the estimate is dominated by a few draws

**Diagnostics and fixes**

- Report $S_{\text{eff}}$ per observation; a small value flags an influential point
- **Pareto-smoothed importance sampling** (Vehtari, Gelman & Gabry, 2017): fit a generalized Pareto distribution to the largest weights, replace them by smoothed values, and use the estimated shape $\hat k$ as a diagnostic ($\hat k>0.7$ unreliable)
- Fall back to actual refitting for the flagged observations only

## Example 5 (Continued)

Normal model, $\theta\sim N(0,10^2)$, 30 observations plus one outlier at $y = 6$. Weight distributions for a typical point and for the outlier.

```{r}
#| fig-width: 10
#| fig-height: 3.8
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
y2 <- c(y, 6); n2 <- length(y2); v <- 1 / (1 / s2p + n2); m <- v * sum(y2)
ths <- rnorm(4000, m, sqrt(v))
for (i in c(1, n2)) { lw <- -dnorm(y2[i], ths, log = TRUE); w <- exp(lw - max(lw)); wt <- w / sum(w)
  plot(sort(wt, decreasing = TRUE)[1:150], type = "h", lwd = 2, col = if (i == 1) "steelblue" else "firebrick",
       xlab = "draw (sorted by weight)", ylab = TeX(r"($\tilde{w}_s$)"),
       main = sprintf("%s (y = %.2f):  S_eff = %.0f of 4000", if (i == 1) "typical point" else "outlier", y2[i], ess(w))) }
```

## Summary

- Importance sampling estimates $E_f[a(\theta)]$ and $C_f = \int f(\theta)\,d\theta$ from draws $\theta_i\sim g$ using weights $w_i = f(\theta_i)/g(\theta_i)$:
$$\widehat{C_f/C_g} = \frac1n\sum_i w_i, \qquad \hat A = \sum_i a(\theta_i)\,\tilde w_i, \quad \tilde w_i = \frac{w_i}{\sum_j w_j}$$
- Good story: a proposal shaped like the target on its support (Example 1) gives near-constant weights and $S_{\text{eff}}\approx n$
- Bad stories: a proposal that misses a mode (Example 2) or part of the support (Example 3) fails **silently**, with perfect-looking weights; a proposal with lighter tails fails **noisily**, with a few huge weights
- $S_{\text{eff}} = 1/\sum_i\tilde w_i^2$ diagnoses only the second kind of failure
- Advanced uses: normalizing-constant ratios (free energy, marginal likelihood) via tempering steps; leave-one-out cross-validation from full-posterior draws with $w_i = 1/p(y_i\mid\theta)$
