---
subtitle: "Optimization for Maximum Likelihood Estimation"
author: "Longhai Li"
format: banner_slides-revealjs
code-fold: true
filters:
  - shinylive
execute:
  echo: true
  fig-width: 7
  fig-height: 4
---

```{r}
#| echo: false
set.seed(812)
```

# Review: Likelihood Theory and MLE

## Likelihood Function

A statistical model specifies the probability (density) of data $D$ given a parameter $\theta$: $f(D \mid \theta)$.

- **As a function of $D$** for fixed $\theta$: the sampling distribution.

- **As a function of $\theta$** for fixed (observed) $D$: the **likelihood function**.
$$
L(\theta; D) = f(D \mid \theta).
$$

The likelihood measures how well each value of $\theta$ *explains* the observed data. For iid data $D = (x_1, \ldots, x_n)$,
$$
L(\theta; D) = \prod_{i=1}^n f(x_i \mid \theta), \qquad \ell(\theta) = \log L(\theta; D) = \sum_{i=1}^n \log f(x_i \mid \theta).
$$

We work with the **log-likelihood** $\ell(\theta)$ to avoid underflow of the product and because sums are easier to differentiate.

## Maximum Likelihood Estimator
$$
\hat\theta_{\mathrm{MLE}} = \arg\max_\theta L(\theta; D) = \arg\max_\theta \ell(\theta).
$$

**Example:** 

$Y_1, Y_2, Y_3 \overset{iid}{\sim} \mathrm{Bern}(\theta)$, observed $y = (1, 1, 0)$.

:::: {.columns}

::: {.column width="50%"}
| $\theta$ | $P(1, 1, 0 \mid \theta) = \theta^2(1-\theta)$ |
|:---:|:---|
| $0$ | $0$ |
| $1/2$ | $1/8 = 0.1250$ |
| $2/3$ | $4/27 = 0.1481$ **(max)** |
| $3/4$ | $9/64 = 0.1406$ |
| $1$ | $0$ |
:::

::: {.column width="50%"}
```{r}
#| echo: false
curve(x^2 * (1 - x), 0, 1, xlab = expression(theta), ylab = "likelihood", lwd = 2); abline(v = 2/3, col = 2, lty = 2)
```
:::

::::

## Example: iid Bernoulli

$P(y_i \mid \theta) = \theta^{y_i}(1-\theta)^{1-y_i}$, so with $n_1 = \sum y_i$ and $n_0 = n - n_1$,
$$
L(\theta) = \theta^{n_1}(1-\theta)^{n_0}, \qquad \ell(\theta) = n_1\log\theta + n_0\log(1-\theta).
$$

**Score function** $S(\theta) = \ell'(\theta)$. The MLE solves the **score equation** $S(\theta) = 0$:
$$
S(\theta) = \frac{n_1}{\theta} - \frac{n_0}{1-\theta} = 0 \iff \hat\theta = \frac{n_1}{n} = \overline y.
$$

**Second derivative** $\ell''(\theta) = -\dfrac{n_1}{\theta^2} - \dfrac{n_0}{(1-\theta)^2} < 0$, so this is a maximum.

## Example: Normal Mean with Known Variance

$X_1, \ldots, X_n \overset{iid}{\sim} N(\mu, \sigma_0^2)$:
$$
\ell(\mu) = -\frac{n}{2}\log(2\pi\sigma_0^2) - \frac{1}{2\sigma_0^2}\sum_{i=1}^n (x_i - \mu)^2
= C - \frac{n}{2\sigma_0^2}(\mu - \overline x)^2,
$$
using $\sum(x_i - \mu)^2 = \sum(x_i - \overline x)^2 + n(\overline x - \mu)^2$.
$$
S(\mu) = \frac{n}{\sigma_0^2}(\overline x - \mu) = 0 \implies \hat\mu = \overline x, \qquad -\ell''(\mu) = \frac{n}{\sigma_0^2}.
$$

- The log-likelihood is exactly **quadratic** in $\mu$; the score is **linear**; the curvature is **constant** and grows with $n$.

- $-\ell''$ measures how sharply the data pin down $\mu$: it is the **Fisher information**.

## Log-likelihood, Score and Curvature

```{r}
#| echo: false
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
x20 <- rnorm(20, 1); x50 <- rnorm(50, 1); mu <- seq(0, 2, length = 200)
ll <- function(mu, x) sapply(mu, function(m) sum(dnorm(x, m, log = TRUE)))
plot(mu, ll(mu, x20) - max(ll(mu, x20)), type = "l", lwd = 2, ylab = expression(l(mu) - max), main = "log-likelihood")
lines(mu, ll(mu, x50) - max(ll(mu, x50)), lwd = 2, col = 2); legend("bottomright", c("n = 20", "n = 50"), col = 1:2, lwd = 2, bty = "n")
plot(mu, 20 * (mean(x20) - mu), type = "l", lwd = 2, ylab = expression(S(mu)), main = "score"); lines(mu, 50 * (mean(x50) - mu), lwd = 2, col = 2); abline(h = 0, lty = 3)
plot(mu, rep(20, 200), type = "l", lwd = 2, ylim = c(0, 60), ylab = expression(-l*minute*minute(mu)), main = "curvature"); lines(mu, rep(50, 200), lwd = 2, col = 2)
```

The score crosses zero where $\ell$ is maximized; larger $n$ makes $\ell$ more peaked, the score steeper, and $-\ell''$ larger.

## Example: Cauchy Location

$X_1, \ldots, X_n \overset{iid}{\sim} \mathrm{Cauchy}(\theta)$, $f(x \mid \theta) = \dfrac{1}{\pi\,[1 + (x-\theta)^2]}$:
$$
\ell(\theta) = -n\log\pi - \sum_{i=1}^n \log\!\big(1 + (x_i - \theta)^2\big), \qquad
S(\theta) = \sum_{i=1}^n \frac{2(x_i - \theta)}{1 + (x_i - \theta)^2}.
$$

The score equation has **no closed-form solution** and may have **several roots**:

```{r}
xc <- rcauchy(8, location = 3)
llc <- function(th, dat) sapply(th, function(t) sum(dcauchy(dat, t, log = TRUE)))
curve(llc(x, xc), -10, 15, lwd = 2, xlab = expression(theta), ylab = expression(l(theta)), main = "Cauchy log-likelihood, n = 8")
```

This is why we need **numerical optimization**.

## Properties of the MLE

Under regularity conditions (satisfied by most models in practice), for iid data with true parameter $\theta_*$:

1. **Consistency**: $\hat\theta_{\mathrm{MLE}} \to \theta_*$ as $n \to \infty$.

2. **Asymptotic normality**:
$$
\hat\theta_{\mathrm{MLE}} \;\dot\sim\; N\!\big(\theta_*,\ [n I(\theta_*)]^{-1}\big), \qquad
I(\theta) = -E_X\!\left[\frac{\partial^2}{\partial\theta\,\partial\theta'}\log f(X \mid \theta)\right],
$$
the **expected Fisher information** per observation ($p \times p$ for a $p$-dimensional $\theta$).

3. **Asymptotic efficiency**: no consistent estimator has smaller asymptotic variance (Cramér–Rao bound).

Note $n I(\theta) = -E_D\big[\ell''(\theta; D)\big]$: the expected curvature of the log-likelihood.

## Observed and Expected Fisher Information

$\theta_*$ is unknown, so the variance $[nI(\theta_*)]^{-1}$ is estimated by plugging in $\hat\theta$. Two versions:

- **Expected information** $n I(\hat\theta)$: requires computing $E_X[\cdot]$ analytically.

- **Observed information** $J(\hat\theta) = -\ell''(\hat\theta; D) = -\sum_{i=1}^n \frac{\partial^2}{\partial\theta\partial\theta'}\log f(x_i \mid \hat\theta)$: the actual curvature at the MLE, no expectation needed.

By the LLN, $J(\hat\theta)/n \approx I(\hat\theta)$. Both give
$$
\widehat{\mathrm{Var}}(\hat\theta) \approx J(\hat\theta)^{-1} \quad\text{or}\quad [nI(\hat\theta)]^{-1}, \qquad \widehat{\mathrm{SE}}(\hat\theta_j) = \sqrt{\big[\widehat{\mathrm{Var}}(\hat\theta)\big]_{jj}} .
$$

**Computational consequence**: the Hessian of $\ell$ at the optimum, which Newton-type optimizers compute anyway, gives the standard errors for free.

# Univariate Optimization

## The Problem

Find $\hat\theta$ maximizing $\ell(\theta)$, $\theta \in \mathbb{R}$. Equivalently, minimize $f(\theta) = -\ell(\theta)$, or find a root of $S(\theta) = \ell'(\theta)$.

Methods differ in **what they need** and **how fast they converge**:

| method | needs | convergence |
|:--|:--|:--|
| fixed-point iteration | a rearrangement $\theta = g(\theta)$ | linear, if at all |
| bisection | $S(\theta)$ and a sign-changing bracket | linear (halves bracket) |
| golden section | $\ell(\theta)$ only, and a bracket | linear |
| **Newton–Raphson** | $S(\theta)$ and $S'(\theta) = \ell''(\theta)$ | **quadratic** near root |
| Fisher scoring | $S(\theta)$ and $I(\theta)$ | linear–quadratic |


## Running Example: Zero-Truncated Poisson {.smaller}

We observe only the **nonzero** counts of a Poisson process ($X = Y \mid Y > 0$, $Y \sim \mathrm{Pois}(\lambda)$):
$$
P(x \mid \lambda) = \frac{e^{-\lambda}\lambda^x}{x!}\cdot\frac{1}{1 - e^{-\lambda}}, \qquad x = 1, 2, \ldots
$$
$$
\ell(\lambda) = -n\lambda + \sum x_i \log\lambda - n\log(1 - e^{-\lambda}) + C
$$
$$
S(\lambda) = \frac{\sum x_i}{\lambda} - \frac{n}{1 - e^{-\lambda}}
$$
$$
S'(\lambda) = -\frac{\sum x_i}{\lambda^2} + \frac{n e^{-\lambda}}{(1 - e^{-\lambda})^2}.
$$

Using $E(X) = \dfrac{\lambda}{1 - e^{-\lambda}}$,
$$
nI(\lambda) = -E[S'(\lambda)] = \frac{n}{\lambda(1 - e^{-\lambda})} - \frac{n e^{-\lambda}}{(1 - e^{-\lambda})^2}.
$$

## Zero-Truncated Poisson in R

```{r}
y <- rpois(200, 1.5); x <- y[y > 0]; n <- length(x); sx <- sum(x)
ll <- function(l) -n * l + sx * log(l) - n * log(1 - exp(-l))
S  <- function(l) sx / l - n / (1 - exp(-l))
dS <- function(l) -sx / l^2 + n * exp(-l) / (1 - exp(-l))^2
nI <- function(l) n / (l * (1 - exp(-l))) - n * exp(-l) / (1 - exp(-l))^2
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
curve(ll, 0.3, 4, lwd = 2, main = "log-likelihood", xlab = expression(lambda))
curve(S, 0.3, 4, lwd = 2, main = "score", xlab = expression(lambda)); abline(h = 0, lty = 3)
c(naive_mean = mean(x), true_lambda = 1.5)
```

The plain sample mean **overestimates** $\lambda$ because the zeros are missing; the MLE corrects for this.

## Fixed-Point (Simple) Iteration

Rearrange $S(\lambda) = 0$ into the form $\lambda = g(\lambda)$:
$$
\frac{\overline x}{\lambda} = \frac{1}{1 - e^{-\lambda}} \iff \lambda = \overline x\,(1 - e^{-\lambda}) =: g(\lambda).
$$

Iterate $\lambda_{k+1} = g(\lambda_k)$. Converges to a fixed point $\lambda^* = g(\lambda^*)$ if $|g'(\lambda^*)| < 1$; the error shrinks by the factor $|g'(\lambda^*)|$ each step (**linear** convergence).

```{r}
fixed_point <- function(g, x0, tol = 1e-8, maxit = 500) {
  for (k in 1:maxit) { x1 <- g(x0); if (abs(x1 - x0) < tol) return(c(root = x1, iter = k)); x0 <- x1 }
  stop("not converged")
}
fixed_point(function(l) mean(x) * (1 - exp(-l)), x0 = mean(x))
```

Advantage: no derivatives. Disadvantage: a suitable $g$ must be found, and convergence can be slow or fail.

## Bisection

If $S$ is continuous and $S(a)$, $S(b)$ have opposite signs, a root lies in $[a, b]$. Evaluate $S$ at the midpoint and keep the half that still brackets the root.

```{r}
bisection <- function(S, a, b, tol = 1e-8) {
  k <- 0
  while (b - a > tol) { m <- (a + b) / 2; if (S(a) * S(m) <= 0) b <- m else a <- m; k <- k + 1 }
  c(root = (a + b) / 2, iter = k)
}
bisection(S, 0.1, 10)
```

- Guaranteed to converge; the bracket halves each step, so $\log_2\big((b-a)/\text{tol}\big) \approx 30$ iterations here.

- Needs only $S$, not $S'$. R's `uniroot()` uses a refined version (Brent's method).

## Golden Section Search

Works on $f(\theta) = -\ell(\theta)$ directly, **without any derivative**: keep a bracket $[a, b]$ containing the minimum of a unimodal $f$ and two interior points $a' < b'$ placed at the golden ratio ($\phi \approx 0.618$).

- If $f(a') < f(b')$ the minimum is in $[a, b']$; otherwise in $[a', b]$.

- The golden ratio placement lets the surviving interior point be **reused**, so each step costs one new evaluation and shrinks the bracket by $0.618$.

```{r}
golden <- function(f, a, b, tol = 1e-8) {
  phi <- (sqrt(5) - 1) / 2; k <- 0
  x1 <- b - phi * (b - a); x2 <- a + phi * (b - a); f1 <- f(x1); f2 <- f(x2)
  while (b - a > tol) {
    if (f1 < f2) { b <- x2; x2 <- x1; f2 <- f1; x1 <- b - phi * (b - a); f1 <- f(x1) }
    else         { a <- x1; x1 <- x2; f1 <- f2; x2 <- a + phi * (b - a); f2 <- f(x2) }
    k <- k + 1 }
  c(root = (a + b) / 2, iter = k)
}
golden(function(l) -ll(l), 0.1, 10)      # R equivalent: optimize(function(l) -ll(l), c(0.1, 10))
```

## Newton–Raphson: Derivation

Approximate the score by its tangent line at the current point $\theta_k$ and solve for where the tangent hits zero:
$$
S(\theta) \approx S(\theta_k) + S'(\theta_k)(\theta - \theta_k) = 0
\quad\Longrightarrow\quad
\theta_{k+1} = \theta_k - \frac{S(\theta_k)}{S'(\theta_k)} = \theta_k - \frac{\ell'(\theta_k)}{\ell''(\theta_k)} .
$$

```{r}
#| echo: false
curve(S, 0.5, 4, lwd = 2, xlab = expression(lambda), ylab = expression(S(lambda)), main = "one Newton step"); abline(h = 0, lty = 3)
l0 <- 3.5; l1 <- l0 - S(l0) / dS(l0)
segments(l0, S(l0), l1, 0, col = 2, lwd = 2); points(c(l0, l1), c(S(l0), 0), pch = 19, col = 2)
text(l0, S(l0), expression(lambda[k]), pos = 4); text(l1, 0, expression(lambda[k+1]), pos = 1)
```

## Newton–Raphson: Interpretation as a Quadratic Model

Equivalently, approximate $\ell$ by its second-order Taylor expansion at $\theta_k$ and maximize that quadratic:
$$
\ell(\theta) \approx \ell(\theta_k) + \ell'(\theta_k)(\theta - \theta_k) + \tfrac12 \ell''(\theta_k)(\theta - \theta_k)^2 .
$$

For a quadratic $\ell(\theta) = -a(\theta - u)^2 + C$ with $a > 0$: $\ell' = -2a(\theta - u)$, $\ell'' = -2a$, so
$$
\theta_{k+1} = \theta_k - \frac{-2a(\theta_k - u)}{-2a} = u \quad\text{in one step}.
$$

- $-\ell''(\theta_k)^{-1}$ is the **perfect step size** for slope $\ell'(\theta_k)$ when $\ell$ is quadratic.

- Log-likelihoods are approximately quadratic near the MLE (this is what asymptotic normality says), so Newton's method is extremely fast there.
- The normal-mean example is exactly quadratic: Newton converges in one step from anywhere.

## Newton–Raphson in R

```{r}
newton <- function(S, dS, x0, tol = 1e-8, maxit = 100, trace = FALSE) {
  for (k in 1:maxit) {
    x1 <- x0 - S(x0) / dS(x0)
    if (trace) cat(sprintf("iter %2d: %.10f\n", k, x1))
    if (!is.finite(x1)) stop("diverged")
    if (abs(x1 - x0) < tol) return(c(root = x1, iter = k))
    x0 <- x1 }
  stop("not converged")
}
newton(S, dS, x0 = mean(x), trace = TRUE)
```

Compare iteration counts: fixed point `r fixed_point(function(l) mean(x) * (1 - exp(-l)), mean(x))["iter"]`, bisection 30, golden 40, Newton just a handful.

## Convergence Rate of Newton–Raphson

Let $e_k = \theta_k - \hat\theta$. Expanding $S(\hat\theta) = 0$ around $\theta_k$:
$$
0 = S(\theta_k) - S'(\theta_k)e_k + \tfrac12 S''(\xi)e_k^2
\quad\Longrightarrow\quad
e_{k+1} = e_k - \frac{S(\theta_k)}{S'(\theta_k)} = \frac{S''(\xi)}{2S'(\theta_k)}\,e_k^2 .
$$

**Quadratic convergence**: the error is squared each step, so the number of correct digits roughly **doubles**. Linear methods (bisection, fixed point) gain a fixed number of digits per step.

```{r}
#| echo: false
lhat <- newton(S, dS, mean(x))["root"]
err_newton <- {e <- c(); x0 <- 3; for (k in 1:6) { x0 <- x0 - S(x0)/dS(x0); e <- c(e, abs(x0 - lhat)) }; e}
err_bisec <- {a <- 0.1; b <- 10; e <- c(); for (k in 1:6) { m <- (a+b)/2; if (S(a)*S(m) <= 0) b <- m else a <- m; e <- c(e, abs((a+b)/2 - lhat)) }; e}
plot(1:6, pmax(err_newton, 1e-16), log = "y", type = "b", pch = 19, ylim = c(1e-16, 10), xlab = "iteration", ylab = "|error| (log scale)")
lines(1:6, err_bisec, type = "b", col = 2, pch = 19); legend("bottomleft", c("Newton", "bisection"), col = 1:2, pch = 19, bty = "n")
```

## When Newton–Raphson Fails {.smaller}

Newton uses only **local** information, so it can misbehave:

- **Wrong curvature**: if $\ell''(\theta_k) > 0$ (a convex region), the step moves toward a *minimum* or diverges.

- **Overshooting**: a flat score ($S'(\theta_k) \approx 0$) gives a huge step.
- **Multiple roots**: it converges to whichever root is nearest the start, possibly a local maximum.

```{r}
Sc  <- function(t) sum(2 * (x_c - t) / (1 + (x_c - t)^2))       # Cauchy score
dSc <- function(t) sum(2 * ((x_c - t)^2 - 1) / (1 + (x_c - t)^2)^2)
x_c <- rcauchy(8, 3)
starts <- c(-5, 0, median(x_c), 8)
sapply(starts, function(s) tryCatch(newton(Sc, dSc, s)["root"], error = function(e) NA))
```

Even the well-behaved truncated Poisson overshoots from a poor start: at $\lambda_0 = 3.5$ the score is flat, the step is huge, and the iterate goes negative.

```{r}
c(step_from_3.5 = 3.5 - S(3.5) / dS(3.5))
tryCatch(newton(S, dS, 3.5), error = function(e) conditionMessage(e))
```

Safeguards: start from a robust estimate (median for Cauchy), **halve the step** until $\ell$ increases, or fall back to bisection/golden section, which only need a bracket.

## Univariate Fisher Scoring

Replace the observed curvature $-\ell''(\theta_k)$ by its expectation $nI(\theta_k)$:
$$
\theta_{k+1} = \theta_k + \frac{S(\theta_k)}{n I(\theta_k)} .
$$

- $nI(\theta) > 0$ always, so the step is always in an ascent direction — more robust than Newton far from the optimum.

- Near the MLE $nI(\hat\theta) \approx -\ell''(\hat\theta)$, so the two methods behave alike.
- Requires the expectation in closed form.

```{r}
fisher_scoring <- function(S, nI, x0, tol = 1e-8, maxit = 100) {
  for (k in 1:maxit) { x1 <- x0 + S(x0) / nI(x0); if (abs(x1 - x0) < tol) return(c(root = x1, iter = k)); x0 <- x1 }
}
rbind(newton = newton(S, dS, 3), fisher = fisher_scoring(S, nI, 3))
```

## Standard Error from the Curvature

At the MLE $\hat\lambda$, the observed information $J(\hat\lambda) = -\ell''(\hat\lambda) = -S'(\hat\lambda)$ and the expected information $nI(\hat\lambda)$ both estimate the precision:

```{r}
lhat <- newton(S, dS, mean(x))["root"]
c(mle = lhat, se_observed = 1 / sqrt(-dS(lhat)), se_expected = 1 / sqrt(nI(lhat)))
lhat + c(-1.96, 1.96) / sqrt(-dS(lhat))    # 95% Wald interval
```

## Numerical Derivatives

When $S$ or $S'$ is tedious to derive, approximate by finite differences:
$$
\ell'(\theta) \approx \frac{\ell(\theta + h) - \ell(\theta - h)}{2h}, \qquad
\ell''(\theta) \approx \frac{\ell(\theta + h) - 2\ell(\theta) + \ell(\theta - h)}{h^2}.
$$

- $h$ too large: truncation error; $h$ too small: roundoff error (large − large). Typical choice $h \approx \sqrt{\epsilon}\,(1 + \lvert\theta
\rvert) \approx 10^{-8}$ for the first derivative, $\epsilon^{1/3}$ for the second.

- R's `nlm()` and `optim(method = "BFGS")` use numerical gradients unless you supply analytic ones; `numDeriv::grad()` and `hessian()` are more accurate.

```{r}
h <- 1e-5
c(analytic = dS(lhat), numeric = (ll(lhat + h) - 2 * ll(lhat) + ll(lhat - h)) / h^2)
```

# Multivariate Newton–Raphson and Fisher Scoring

## Gradient and Hessian

For $\theta = (\theta_1, \ldots, \theta_p)'$, the **gradient** (score vector) and **Hessian** are
$$
\nabla\ell(\theta) = \begin{pmatrix}\partial\ell/\partial\theta_1 \\ \vdots \\ \partial\ell/\partial\theta_p\end{pmatrix}, \qquad
\nabla^2\ell(\theta) = \left(\frac{\partial^2\ell}{\partial\theta_i\,\partial\theta_j}\right)_{p\times p} .
$$


- $\nabla\ell$ points in the direction of **steepest ascent** and is perpendicular to the contour through $\theta$.

- $\nabla^2\ell$ describes the local curvature. At a local maximum it is negative semidefinite; if it is negative definite, the maximum is strict and locally unique. Near a maximum, $\ell$ is well approximated by a quadratic, so the Hessian determines the shape of the contours.


## Gradient and Hessian: Illustration

Quadratic: $\ell(\theta) = -\tfrac12(\theta - u)'A(\theta - u)$, $A$ positive definite, $\nabla\ell = -A(\theta - u)$, $\nabla^2\ell = -A$ (constant). Contours are ellipses centred at $u$. Write $A = Q\Lambda Q'$:

:::: {.columns}

::: {.column width="55%"}

- axes point along the eigenvectors of $Q$;
- half-length along axis $j$ is $\propto 1/\sqrt{\lambda_j}$, so a large eigenvalue means high curvature and a short, well-determined direction;
- $\lambda_{\max}/\lambda_{\min}$ measures elongation; tilted ellipses indicate correlated parameters.
- Gradient arrows (green) are perpendicular to contours and point uphill, but not toward $u$ unless the ellipse is a circle.
- Near its maximum any smooth $\ell$ looks like this, with $A = -\nabla^2\ell(\hat\theta)$.
:::

::: {.column width="45%"}
```{r}
#| echo: false
#| fig-width: 7
#| fig-height: 7
A <- matrix(c(2, 1.2, 1.2, 1), 2)
s <- seq(-2, 2, length.out = 80); g <- expand.grid(x = s, y = s)
z <- matrix(apply(g, 1, function(v) -0.5 * drop(t(v) %*% A %*% v)), 80)
par(pty = "s", mar = c(4, 4, 1, 1))
contour(s, s, z, nlevels = 12, xlab = expression(theta[1]), ylab = expression(theta[2]))
# gradient arrows
pts <- rbind(c(1, 1), c(-1.5, 0.5), c(0.5, -1.5))
for (i in 1:3) { gr <- -A %*% pts[i, ]
  arrows(pts[i,1], pts[i,2], pts[i,1] + 0.3*gr[1], pts[i,2] + 0.3*gr[2],
         col = 3, lwd = 2, length = 0.1) }
# principal axes with eigenvalue labels: length 1/sqrt(lambda) on each side
e <- eigen(A); k <- 1.6
for (j in 1:2) { v <- e$vectors[, j] * k / sqrt(e$values[j])
  segments(-v[1], -v[2], v[1], v[2], col = "firebrick", lwd = 2, lty = 2)
  text(v[1], v[2], bquote(lambda[.(j)] == .(round(e$values[j], 2))),
       col = "firebrick", pos = if (j == 1) 4 else 2, cex = 0.9) }
points(0, 0, pch = 19)
```
:::

::::

## Multivariate Newton–Raphson

Second-order Taylor expansion at $\theta_k$, maximized exactly:
$$
\ell(\theta) \approx \ell(\theta_k) + \nabla\ell(\theta_k)'(\theta - \theta_k) + \tfrac12(\theta - \theta_k)'\nabla^2\ell(\theta_k)(\theta - \theta_k)
$$
$$
\Longrightarrow\quad
\underbrace{\theta_{k+1}}_{p\times1} = \theta_k - \underbrace{\big[\nabla^2\ell(\theta_k)\big]^{-1}}_{p\times p}\underbrace{\nabla\ell(\theta_k)}_{p\times1}.
$$

Compare with the univariate $\theta_{k+1} = \theta_k - \ell'(\theta_k)/\ell''(\theta_k)$: division becomes multiplication by the inverse Hessian.

**Fisher scoring** replaces $-\nabla^2\ell(\theta_k)$ by the expected information $nI(\theta_k)$:
$$
\theta_{k+1} = \theta_k + \big[nI(\theta_k)\big]^{-1}\nabla\ell(\theta_k).
$$

Each iteration costs one $p \times p$ linear solve — $O(p^3)$ — fine for $p$ in the hundreds, expensive for $p$ in the millions.

## Example: Logistic Regression {.smaller}

$y_i \mid x_i \sim \mathrm{Bern}(\pi_i)$ with $\pi_i = \dfrac{e^{\eta_i}}{1 + e^{\eta_i}} = \dfrac{1}{1 + e^{-\eta_i}}$, $\eta_i = x_i'\beta$, $x_i = (1, x_{i1}, \ldots, x_{ip})'$.
$$
L(\beta) = \prod_{i=1}^n \pi_i^{y_i}(1-\pi_i)^{1-y_i}
= \prod_{i=1}^n \frac{e^{y_i\eta_i}}{1 + e^{\eta_i}}
$$
$$
\ell(\beta) = \sum_{i=1}^n \big[y_i\,x_i'\beta - \log(1 + e^{x_i'\beta})\big].
$$

Using $\dfrac{\partial}{\partial\beta}\log(1 + e^{x_i'\beta}) = \pi_i x_i$ and $\dfrac{\partial\pi_i}{\partial\beta} = \pi_i(1-\pi_i)x_i$:
$$
\nabla\ell(\beta) = \sum_{i=1}^n (y_i - \pi_i)\,x_i = X'(y - \pi), \qquad
$$
$$
\nabla^2\ell(\beta) = -\sum_{i=1}^n \pi_i(1-\pi_i)\,x_i x_i' = -X'WX, \text{with}~W = \mathrm{diag}\big(\pi_i(1-\pi_i)\big)
$$
.

## Logistic Regression: Newton = Fisher Scoring = IRLS

- $-\nabla^2\ell = X'WX$ is positive semi-definite, so $\ell$ is **concave**: any stationary point is the global maximum.

- The Hessian does not involve $y$, so its expectation equals itself: **Newton–Raphson and Fisher scoring coincide**.
- The update
$$
\beta_{k+1} = \beta_k + (X'W_kX)^{-1}X'(y - \pi_k) = (X'W_kX)^{-1}X'W_k z_k,
$$ 
$$
z_k = X\beta_k + W_k^{-1}(y - \pi_k),
$$
is a weighted least squares fit of the *working response* $z_k$: **iteratively reweighted least squares (IRLS)**, which is how `glm()` fits all generalized linear models.

- Caveat: if the classes are perfectly separable, $\ell$ increases forever and the MLE does not exist ($\lvert\hat\beta\rvert \to \infty$).

## Logistic Regression in R

```{r}
n <- 500; X <- cbind(1, rnorm(n), rnorm(n)); beta_true <- c(-0.5, 1, -2)
y <- rbinom(n, 1, plogis(X %*% beta_true))
loglik <- function(b) sum(y * (X %*% b) - log1p(exp(X %*% b)))
grad   <- function(b) crossprod(X, y - plogis(X %*% b))
hess   <- function(b) { p <- plogis(X %*% b); -crossprod(X, X * drop(p * (1 - p))) }

newton_mv <- function(grad, hess, b0, tol = 1e-8, maxit = 50) {
  path <- drop(b0)
  for (k in 1:maxit) {
    b1 <- b0 - solve(hess(b0), grad(b0)); path <- rbind(path, drop(b1))
    if (max(abs(b1 - b0)) < tol)
      return(list(est = drop(b1), iter = k, se = sqrt(diag(solve(-hess(b1)))), path = path))
    b0 <- b1 }
}
fit <- newton_mv(grad, hess, rep(0, 3)); fit$iter
rbind(newton = fit$est, glm = coef(glm(y ~ X[, -1], family = binomial)))
rbind(newton = fit$se, glm = sqrt(diag(vcov(glm(y ~ X[, -1], family = binomial)))))
```

## Variance of the MLE $\hat\beta$
$$
\widehat{\mathrm{Var}}(\hat\beta) = \big[-\nabla^2\ell(\hat\beta)\big]^{-1} = (X'\hat WX)^{-1}
= \begin{pmatrix}\hat\sigma_{00} & \cdots & \hat\sigma_{0p} \\ \vdots & \ddots & \vdots \\ \hat\sigma_{p0} & \cdots & \hat\sigma_{pp}\end{pmatrix},
$$
$$
\widehat{\mathrm{SE}}(\hat\beta_j) = \sqrt{\hat\sigma_{jj}} .
$$

Wald tests and confidence intervals follow directly:
$$
z_j = \frac{\hat\beta_j}{\widehat{\mathrm{SE}}(\hat\beta_j)} \;\dot\sim\; N(0,1) \text{ under } H_0: \beta_j = 0, \qquad \hat\beta_j \pm 1.96\,\widehat{\mathrm{SE}}(\hat\beta_j).
$$

This is the `summary(glm(...))` table. The variance comes from the same Hessian that drove the Newton iterations.

## The Need to Robustify Newton–Raphson

The Newton step $d_k = -H_k^{-1} g_k$, with $g_k = \nabla f(x_k)$ and
$H_k = \nabla^2 f(x_k)$, solves the local quadratic model exactly. Two things
can go wrong away from the solution:

- **The direction may not descend.** $g_k^\top d_k = -g_k^\top H_k^{-1} g_k < 0$
  requires $H_k \succ 0$. Where $f$ is not convex, $H_k$ may be indefinite or
  singular and $d_k$ can point uphill.
- **The unit step may overshoot.** Even with $H_k \succ 0$, the quadratic model
  is trustworthy only near $x_k$; $\alpha_k = 1$ can increase $f$.

The two failures are independent and need separate remedies: one fixes the
**direction**, the other the **length**.

## Robustification of the Hessian

Replace $H_k$ by a positive definite $\tilde H_k$ and solve
$\tilde H_k d_k = -g_k$.

**Shift by a multiple of the identity.** With $\lambda_{\min}$ the smallest
eigenvalue of $H_k$,
$$
\tilde H_k = H_k + \tau I, \qquad
\tau = \max\left(0,\; \delta - \lambda_{\min}\right), \qquad
\delta = 10^{-3}\lambda_{\max} .
$$
Eigenvectors are unchanged and every eigenvalue is shifted by $\tau$. As
$\tau \to \infty$ the direction rotates toward $-g_k$, so the method
interpolates between Newton and steepest descent.

**Spectral modification.** With $H_k = V \Lambda V^\top$, replace
$\lambda_i$ by $\max(|\lambda_i|, \delta)$. This reverses negative-curvature
directions rather than damping them, so the step moves *away* from a saddle.

Where $H_k \succ 0$ already, $\tau = 0$ and $\tilde H_k = H_k$: the refinement
is inactive near a well-behaved minimum.

## Step length: backtracking and line search

Given a descent direction, choose $\alpha_k$ by an inexact search along
$\phi(\alpha) = f(x_k + \alpha d_k)$.

**Armijo (sufficient decrease).**
$$
f(x_k + \alpha d_k) \le f(x_k) + c_1 \alpha\, g_k^\top d_k,
\qquad c_1 = 10^{-4}.
$$

**Backtracking.** Start at $\alpha = 1$ and set $\alpha \leftarrow \rho\alpha$
with $\rho = \tfrac12$ until Armijo holds. Simple, and it accepts $\alpha_k = 1$
whenever the full Newton step is acceptable.

**Curvature (Wolfe) condition.**
$$
g(x_k + \alpha d_k)^\top d_k \ \ge\ c_2\, g_k^\top d_k,
\qquad c_2 = 0.9,
$$
which rules out steps that are too short. Backtracking alone cannot enforce it,
since it never lengthens the trial step. The condition matters for conjugate
gradient and is what keeps $s_k^\top y_k > 0$, hence the BFGS update well
defined.

## The safeguarded algorithm

::: {.callout-note appearance="minimal"}
At $x_k$: compute $g_k$, $H_k$; form $\tilde H_k \succ 0$; solve
$\tilde H_k d_k = -g_k$; find $\alpha_k$ from $\alpha_0 = 1$ by line search;
set $x_{k+1} = x_k + \alpha_k d_k$.
:::

Both safeguards are needed, and both switch themselves off:

| | near a minimum | far away / indefinite |
|---|---|---|
| $\tau$ | $0$ | $> 0$, direction rotates toward $-g_k$ |
| $\alpha_k$ | $1$ | $< 1$ (or $> 1$ under Wolfe) |

Quadratic convergence is therefore retained locally, while sufficient decrease
at every iteration gives convergence from remote starting points. This is
**safeguarded Newton**, not "Newton–Raphson" in the textbook sense — and not
what `nlm()` runs, which globalizes with a trust region.

# Gradient-based Methods

## What If the Hessian Is Unavailable or Too Big?

| method | uses | per-iteration cost | typical use |
|:--|:--|:--|:--|
| Nelder–Mead | $f$ only | $O(p)$ evaluations | small $p$, non-smooth or noisy $f$ |
| gradient descent + line search | $\nabla f$ | $O(p)$ + line search | large $p$ |
| conjugate gradient | $\nabla f$ | $O(p)$ + line search | large $p$, quadratic-like $f$ |
| quasi-Newton (BFGS, L-BFGS) | $\nabla f$, builds $\approx$ Hessian | $O(p^2)$ / $O(mp)$ | default general-purpose |
| Newton / Fisher scoring | $\nabla f$, $\nabla^2 f$ | $O(p^3)$ | moderate $p$, gives SEs |

All are available through `optim(par, fn, gr, method = ...)` in R. Convention below: **minimize** $f(\theta) = -\ell(\theta)$.


## Gradient Descent with Line Search

Move in the direction of steepest descent $d_k = -\nabla f(\theta_k)$, choosing the step length $\alpha_k$ by a **line search**:
$$
\theta_{k+1} = \theta_k + \alpha_k d_k, \qquad \alpha_k = \arg\min_{\alpha > 0} f(\theta_k + \alpha d_k) .
$$

- Exact line search: a univariate minimization (golden section / `optimize`) along the ray.

- Inexact (**backtracking, Armijo**): start with $\alpha = 1$ and halve until $f(\theta_k + \alpha d_k) \le f(\theta_k) + c\,\alpha\,\nabla f(\theta_k)'d_k$ (a "sufficient decrease").

```{r}
grad_descent <- function(f, gr, x0, maxit = 1000, tol = 1e-8) {
  path <- x0
  for (k in 1:maxit) {
    d <- -gr(x0); a <- optimize(function(a) f(x0 + a * d), c(0, 10))$minimum   # exact line search
    x1 <- x0 + a * d; path <- rbind(path, x1)
    if (sqrt(sum((x1 - x0)^2)) < tol) break
    x0 <- x1 }
  list(par = x1, iter = k, path = path)
}
```

## Why Gradient Descent Zig-Zags

With exact line search, consecutive directions are **orthogonal** ($\nabla f(\theta_{k+1})'d_k = 0$), so on elongated contours the path zig-zags across the valley:

```{r}
#| echo: false
fq <- function(v) 0.5 * drop(t(v) %*% A %*% v); gq <- function(v) drop(A %*% v)
gd <- grad_descent(fq, gq, c(2, -1.8), maxit = 30)
par(pty = "s", mar = c(4, 4, 2, 1)); contour(seq(-2, 2, l = 80), seq(-2, 2, l = 80), -z, nlevels = 15, main = sprintf("gradient descent, %d iterations", gd$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6)
```

The number of iterations grows with the **condition number** of the Hessian (ratio of largest to smallest eigenvalue).

## Conjugate Gradient

Fix the zig-zag by making each new direction **conjugate** to the previous ones with respect to the Hessian $Q$: $d_i'Qd_j = 0$ for $i \ne j$. For a quadratic $f(\theta) = \tfrac12\theta'Q\theta - b'\theta$, minimizing along $p$ such directions reaches the exact minimum in **at most $p$ steps**.

The nonlinear version builds directions from gradients only:
$$
d_0 = -g_0, \qquad d_{k+1} = -g_{k+1} + \beta_k d_k, \qquad
$$
$$
\beta_k^{FR} = \frac{g_{k+1}'g_{k+1}}{g_k'g_k}\ \text{(Fletcher–Reeves)}, \quad
\beta_k^{PR} = \frac{g_{k+1}'(g_{k+1} - g_k)}{g_k'g_k}\ \text{(Polak–Ribière)},
$$
with $\theta_{k+1} = \theta_k + \alpha_k d_k$ from a line search and $g_k = \nabla f(\theta_k)$.

- Memory: only two vectors. No matrix ever formed. Ideal for very large $p$.

- Restart with $d = -g$ every $p$ iterations, since conjugacy degrades on non-quadratic $f$.

## Conjugate Gradient in R

```{r}
conj_grad <- function(f, gr, x0, maxit = 1000, tol = 1e-8, restart = length(x0)) {
  g0 <- gr(x0); d <- -g0; path <- x0
  for (k in 1:maxit) {
    a <- optimize(function(a) f(x0 + a * d), c(0, 10))$minimum
    x1 <- x0 + a * d; g1 <- gr(x1); path <- rbind(path, x1)
    if (sqrt(sum(g1^2)) < tol) break
    beta <- if (k %% restart == 0) 0 else max(0, sum(g1 * (g1 - g0)) / sum(g0^2))   # Polak-Ribiere+
    d <- -g1 + beta * d; x0 <- x1; g0 <- g1 }
  list(par = x1, iter = k, path = path)
}
cg <- conj_grad(fq, gq, c(2, -1.8))
par(pty = "s", mar = c(4, 4, 2, 1)); contour(seq(-2, 2, l = 80), seq(-2, 2, l = 80), -z, nlevels = 15, main = sprintf("conjugate gradient, %d iterations", cg$iter))
lines(gd$path, type = "b", col = 2, pch = 19, cex = 0.6); lines(cg$path, type = "b", col = 4, pch = 19, lwd = 2)
legend("topleft", c("gradient descent", "conjugate gradient"), col = c(2, 4), lwd = 2, bty = "n")
```

## Quasi-Newton: BFGS

Between gradient methods and Newton: build an approximation $B_k \approx \nabla^2 f$ from successive gradient differences $s_k = \theta_{k+1} - \theta_k$, $y_k = g_{k+1} - g_k$, via a rank-two update that keeps $B_k$ positive definite. Steps are $d_k = -B_k^{-1}g_k$ with a line search.

```{r}
negll <- function(b) -loglik(b); neggr <- function(b) -drop(grad(b))
fits <- list(NM = optim(rep(0, 3), negll, method = "Nelder-Mead"),
             CG = optim(rep(0, 3), negll, neggr, method = "CG"),
             BFGS = optim(rep(0, 3), negll, neggr, method = "BFGS", hessian = TRUE))
t(sapply(fits, function(f) c(f$par, f_evals = f$counts[1], g_evals = f$counts[2])))
sqrt(diag(solve(fits$BFGS$hessian)))     # SEs from the numerically computed Hessian at the optimum
```

`optim(..., hessian = TRUE)` returns a finite-difference Hessian at the solution, so standard errors are available even for gradient-only methods.

## Animation for Gradient-based Methods

```{shinylive-r}
#| standalone: true
#| viewerHeight: 820

library(shiny)

## ---------------------------------------------------------------- objectives
mk <- function(label, fv, xlim, ylim, start, xstar) {
  force(fv)
  list(label = label, fv = fv, f = function(p) fv(p[1], p[2]),
       xlim = xlim, ylim = ylim, start = start,
       xstar = matrix(xstar, ncol = 2))
}

FNS <- list(
  quad = mk(
    "(a) Well-conditioned quadratic",
    function(x, y) 0.5 * (2 * (x - 1)^2 + 1.6 * (x - 1) * (y - 1) +
                          2 * (y - 1)^2),
    c(-3, 4), c(-3, 4), c(-2, 3), c(1, 1)
  ),
  banana = mk(
    "(b) Rosenbrock banana (narrow curved ridge)",
    function(x, y) (1 - x)^2 + 100 * (y - x^2)^2,
    c(-2, 2), c(-1, 3), c(-1.2, 1), c(1, 1)
  ),
  mix = mk(
    "(c) Highly correlated, two modes",
    function(x, y) {
      rho <- 0.9; den <- 1 - rho^2
      q <- function(u, v) (u^2 - 2 * rho * u * v + v^2) / den
      -log(exp(-0.5 * q(x - 1.2, y - 1.2)) +
           0.6 * exp(-0.5 * q(x + 1.5, y + 1.5) / 1.5) + 1e-12)
    },
    c(-4, 4), c(-4, 4), c(-3, -0.5), c(1.2, 1.2)
  ),
  himmel = mk(
    "(d) Himmelblau (four global minima)",
    function(x, y) (x^2 + y - 11)^2 + (x + y^2 - 7)^2,
    c(-5, 5), c(-5, 5), c(-4, 4),
    rbind(c(3, 2), c(-2.805118, 3.131312),
          c(-3.779310, -3.283186), c(3.584428, -1.848126))
  ),
  beale = mk(
    "(e) Beale (flat plateau, sharp valley)",
    function(x, y) (1.5 - x + x * y)^2 + (2.25 - x + x * y^2)^2 +
                   (2.625 - x + x * y^3)^2,
    c(-4.5, 4.5), c(-4.5, 4.5), c(-1, 1), c(3, 0.5)
  ),
  rastrigin = mk(
    "(f) Rastrigin (many local minima)",
    function(x, y) 20 + x^2 - 10 * cos(2 * pi * x) +
                        y^2 - 10 * cos(2 * pi * y),
    c(-5.12, 5.12), c(-5.12, 5.12), c(-3.1, 4.2), c(0, 0)
  )
)

## ------------------------------------------------- numerical derivatives
num_grad <- function(f, x) {
  h <- 1e-5 * (1 + abs(x))
  g <- numeric(2)
  for (i in 1:2) {
    e <- numeric(2); e[i] <- h[i]
    g[i] <- (f(x + e) - f(x - e)) / (2 * h[i])
  }
  g
}

num_hess <- function(f, x) {
  h <- 1e-4 * (1 + abs(x))
  H <- matrix(0, 2, 2)
  for (i in 1:2) for (j in i:2) {
    ei <- numeric(2); ei[i] <- h[i]
    ej <- numeric(2); ej[j] <- h[j]
    H[i, j] <- (f(x + ei + ej) - f(x + ei - ej) -
                f(x - ei + ej) + f(x - ei - ej)) / (4 * h[i] * h[j])
    H[j, i] <- H[i, j]
  }
  H
}

## --------------------------------------------------- Wolfe line search
line_search <- function(f, g, x, d, a0 = 1, c1 = 1e-4, c2 = 0.9) {
  f0 <- f(x); gd0 <- sum(g * d)
  if (!is.finite(gd0) || gd0 >= 0) return(list(alpha = 0, ok = FALSE))
  lo <- 0; hi <- Inf; a <- a0
  for (it in 1:60) {
    fa <- f(x + a * d)
    if (!is.finite(fa) || fa > f0 + c1 * a * gd0) {
      hi <- a; a <- 0.5 * (lo + hi)
    } else {
      ga <- num_grad(f, x + a * d)
      if (sum(ga * d) < c2 * gd0) {
        lo <- a
        a <- if (is.finite(hi)) 0.5 * (lo + hi) else min(2 * a, 1e4)
      } else {
        return(list(alpha = a, ok = TRUE))
      }
    }
    if (a < 1e-14) return(list(alpha = 0, ok = FALSE))
  }
  list(alpha = a, ok = (a > 0))
}

## ------------------------------------------------------------- algorithms
run_alg <- function(alg, FN, x0, maxit = 60, tol = 1e-6) {
  f <- FN$f
  X <- matrix(NA_real_, maxit + 1, 2); G <- matrix(NA_real_, maxit + 1, 2)
  D <- matrix(NA_real_, maxit + 1, 2); Fv <- rep(NA_real_, maxit + 1)
  AL <- rep(NA_real_, maxit + 1)
  x <- x0; B <- diag(2)
  g_prev <- NULL; d_prev <- NULL; s_prev <- NULL
  n <- 0; note <- "maximum number of iterations reached"

  for (k in seq_len(maxit)) {
    g <- num_grad(f, x); gn <- sqrt(sum(g^2))
    X[k, ] <- x; G[k, ] <- g; Fv[k] <- f(x); n <- k
    if (!is.finite(gn)) { note <- "diverged"; break }
    if (gn < tol) { note <- "converged (||grad|| < 1e-6)"; break }

    if (alg == "nr") {
      H <- num_hess(f, x); H <- (H + t(H)) / 2
      ev <- eigen(H, symmetric = TRUE, only.values = TRUE)$values
      delta <- 1e-3 * max(1, max(abs(ev)))
      tau <- max(0, delta - min(ev))
      d <- tryCatch(solve(H + tau * diag(2), -g), error = function(e) -g)
    } else if (alg == "gd") {
      d <- -g
    } else if (alg == "cg") {
      if (is.null(d_prev)) {
        d <- -g
      } else {
        y <- g - g_prev
        beta <- max(0, sum(g * y) / max(sum(g_prev^2), 1e-300))   # PR+
        d <- -g + beta * d_prev
      }
      if (sum(g * d) > -1e-12) d <- -g                            # restart
    } else {                                                      # BFGS
      if (!is.null(s_prev)) {
        y <- g - g_prev; sy <- sum(s_prev * y)
        if (sy > 1e-10) {
          rho <- 1 / sy
          V <- diag(2) - rho * outer(s_prev, y)
          B <- V %*% B %*% t(V) + rho * outer(s_prev, s_prev)
        }
      }
      d <- -as.vector(B %*% g)
      if (sum(g * d) > -1e-12) { B <- diag(2); d <- -g }
    }

    ls <- line_search(f, g, x, d)
    D[k, ] <- d; AL[k] <- ls$alpha
    if (ls$alpha <= 0) { note <- "line search failed (flat or non-descent)"; break }

    s_prev <- ls$alpha * d; g_prev <- g; d_prev <- d
    x <- x + s_prev
  }

  list(X = X[1:n, , drop = FALSE], G = G[1:n, , drop = FALSE],
       D = D[1:n, , drop = FALSE], F = Fv[1:n], A = AL[1:n],
       n = n, note = note)
}

ALGS <- c("Safeguarded Newton (modified Hessian + line search)" = "nr",
          "Gradient descent + line search"    = "gd",
          "Conjugate gradient (PR+)"          = "cg",
          "BFGS (quasi-Newton)"               = "bfgs")

## ---------------------------- widen the data box to the device aspect ratio
expand_box <- function(xlim, ylim, pin) {
  if (length(pin) != 2 || any(!is.finite(pin)) || any(pin <= 0))
    return(list(xlim = xlim, ylim = ylim))
  r <- pin[1] / pin[2]
  wd <- diff(xlim); ht <- diff(ylim)
  if (wd / ht < r) {
    w <- ht * r; m <- mean(xlim); xlim <- m + c(-0.5, 0.5) * w
  } else {
    h <- wd / r; m <- mean(ylim); ylim <- m + c(-0.5, 0.5) * h
  }
  list(xlim = xlim, ylim = ylim)
}

## --------------------------------------------------------------------- UI
ui <- fluidPage(
  fluidRow(
    ## ---- column 1: controls -------------------------------------------
    column(
      3,
      selectInput("alg", "Algorithm", ALGS),
      selectInput("fn", "Objective function",
                  setNames(names(FNS), sapply(FNS, `[[`, "label"))),
      sliderInput("k", "Iteration k", min = 0, max = 1, value = 0, step = 1),
      checkboxInput("showpath", "Show the whole path", TRUE),
      htmlOutput("info")
    ),
    ## ---- column 2: playback -------------------------------------------
    column(
      2,
      div(
        style = "margin-top:26px;",
        actionButton("play", "Play", class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("step", "Next", class = "btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("back", "Prev", class = "btn-lg",
                     style = "width:100%;"),
        hr(),
        checkboxInput("autoplay", "Autoplay on Changing Initial", TRUE),
        actionButton("reset", "Reset initial",
                     style = "width:100%; margin-bottom:8px;"),
        helpText("Click inside the contour plot to choose a new initial value.")
      )
    ),
    ## ---- column 3: plots ----------------------------------------------
    column(
      7,
      plotOutput("contour", click = "click", height = "540px"),
      plotOutput("conv", height = "190px")
    )
  )
)

## ----------------------------------------------------------------- server
server <- function(input, output, session) {

  FN <- reactive(FNS[[input$fn]])
  start <- reactiveVal(FNS[[1]]$start)
  playing <- reactiveVal(FALSE)

  observeEvent(input$fn, start(FNS[[input$fn]]$start))
  observeEvent(input$reset, start(FN()$start))
  observeEvent(input$click, {
    p <- c(input$click$x, input$click$y)
    if (all(is.finite(p))) start(p)
  })

  path <- reactive({
    req(start())
    run_alg(input$alg, FN(), start())
  })

  observeEvent(path(), {
    updateSliderInput(session, "k", max = max(1, path()$n - 1), value = 0)
    playing(isTRUE(input$autoplay) && path()$n > 1)
  })

  observeEvent(input$play, {
    if (isTRUE(playing())) {
      playing(FALSE)
    } else {
      if (as.integer(input$k) >= path()$n - 1)
        updateSliderInput(session, "k", value = 0)
      playing(TRUE)
    }
  })

  observeEvent(playing(), {
    updateActionButton(session, "play",
                       label = if (isTRUE(playing())) "Pause" else "Play")
  })

  observe({
    if (!isTRUE(playing())) return()
    kmax <- isolate(path()$n) - 1
    k <- isolate(as.integer(input$k))
    if (k >= kmax) { playing(FALSE); return() }
    invalidateLater(900, session)
    updateSliderInput(session, "k", value = k + 1)
  })

  observeEvent(input$step, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = min(as.integer(input$k) + 1, max(0, path()$n - 1)))
  })

  observeEvent(input$back, {
    playing(FALSE)
    updateSliderInput(session, "k", value = max(as.integer(input$k) - 1, 0))
  })

  output$contour <- renderPlot({
    FNc <- FN(); P <- path()
    k <- min(as.integer(input$k), P$n - 1) + 1L

    par(mar = c(4, 4, 2.5, 1))
    plot.new()
    bx <- expand_box(FNc$xlim, FNc$ylim, par("pin"))
    plot.window(xlim = bx$xlim, ylim = bx$ylim, xaxs = "i", yaxs = "i")

    xs <- seq(bx$xlim[1], bx$xlim[2], length.out = 180)
    ys <- seq(bx$ylim[1], bx$ylim[2], length.out = 180)
    z  <- outer(xs, ys, FNc$fv)
    zf <- z[is.finite(z)]
    lv <- unique(quantile(zf, probs = seq(0, 1, length.out = 30)^1.7))
    zr <- matrix(rank(z, na.last = "keep", ties.method = "average"),
                 nrow = length(xs))

    image(xs, ys, zr, add = TRUE,
          col = colorRampPalette(c("#f7fbff", "#9ec4e3"))(128))
    contour(xs, ys, z, levels = lv, add = TRUE,
            col = "grey45", drawlabels = FALSE)
    axis(1); axis(2); box()
    title(main = paste0(names(ALGS)[ALGS == input$alg], "  |  ", FNc$label),
          xlab = expression(x[1]), ylab = expression(x[2]), cex.main = 1.0)

    points(FNc$xstar[, 1], FNc$xstar[, 2], pch = 8, col = "red",
           cex = 1.3, lwd = 2)

    if (isTRUE(input$showpath) && k > 1)
      lines(P$X[1:k, 1], P$X[1:k, 2], col = "grey25", lty = 2)
    points(P$X[1:k, 1], P$X[1:k, 2], pch = 20, col = "grey35", cex = 0.9)

    xk <- P$X[k, ]; gk <- P$G[k, ]; gn <- sqrt(sum(gk^2))
    L <- 0.09 * diff(bx$xlim)                      # fixed display length
    if (gn > 0) {
      u <- -gk / gn
      arrows(xk[1], xk[2], xk[1] + L * u[1], xk[2] + L * u[2],
             col = "grey45", lwd = 3, length = 0.10)
      arrows(xk[1], xk[2], xk[1] - gk[1], xk[2] - gk[2],
             col = "#E8820C", lwd = 3, length = 0.13)
    }
    if (k < P$n) {
      arrows(xk[1], xk[2], P$X[k + 1, 1], P$X[k + 1, 2],
             col = "#2E8B57", lwd = 3, length = 0.13)
      points(P$X[k + 1, 1], P$X[k + 1, 2], pch = 21, bg = "white",
             col = "#2E8B57", cex = 1.2)
    }
    points(xk[1], xk[2], pch = 19, col = "black", cex = 1.3)

    legend("topleft", lwd = 3, cex = 0.9, bg = "#ffffffcc", box.col = NA,
           col = c("grey45", "#E8820C", "#2E8B57"),
           legend = c(expression(-nabla * f(x[k])~"(direction only)"),
                      expression(-nabla * f(x[k])~"(actual length)"),
                      expression(alpha[k] * d[k]~"(step taken)")))
  })

  output$conv <- renderPlot({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    gn <- sqrt(rowSums(P$G^2))
    gn[!is.finite(gn) | gn <= 0] <- NA
    par(mar = c(4, 4.5, 1.5, 1))
    plot(seq_len(P$n) - 1, gn, type = "b", pch = 20, log = "y",
         xlab = "iteration k", ylab = expression("||" * nabla * f * "||"),
         col = "grey40")
    points(k - 1, gn[k], pch = 19, col = "#E8820C", cex = 1.6)
  })

  output$info <- renderUI({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    xk <- P$X[k, ]; gk <- P$G[k, ]; gn <- sqrt(sum(gk^2))
    ang <- NA
    if (k < P$n) {
      sk <- P$X[k + 1, ] - xk
      cs <- sum(-gk * sk) / (gn * sqrt(sum(sk^2)))
      ang <- acos(max(-1, min(1, cs))) * 180 / pi
    }
    fmt <- function(v, d = 4) formatC(v, format = "g", digits = d)
    HTML(paste0(
      "<hr><b>iteration</b> ", k - 1, " of ", P$n - 1, "<br>",
      "<b>x<sub>k</sub></b> = (", fmt(xk[1]), ", ", fmt(xk[2]), ")<br>",
      "<b>f(x<sub>k</sub>)</b> = ", fmt(P$F[k], 6), "<br>",
      "<b>||grad||</b> = ", fmt(gn), "<br>",
      "<b>step size &alpha;<sub>k</sub></b> = ",
      if (k < P$n) fmt(P$A[k]) else "&ndash;", "<br>",
      "<b>angle(grey, green)</b> = ",
      if (is.na(ang)) "&ndash;" else paste0(fmt(ang, 3), "&deg;"),
      "<br><br><i>", P$note, "</i>"
    ))
  })
}

shinyApp(ui, server)
```



# Nelder–Mead (Simplex Method)

Maintains a **simplex** of $p+1$ points (a triangle in 2D) and moves it downhill using only function values:

1. Order the vertices $f(\theta_{(1)}) \le \cdots \le f(\theta_{(p+1)})$; let $\overline\theta$ be the centroid of the best $p$.

2. **Reflect** the worst vertex through $\overline\theta$; if the reflected point is the new best, try **expanding** further.
3. If reflection is poor, **contract** toward $\overline\theta$; if that also fails, **shrink** all vertices toward the best.

```{r}
rosen <- function(v) (1 - v[1])^2 + 100 * (v[2] - v[1]^2)^2       # Rosenbrock "banana", min at (1, 1)
nm <- optim(c(-1.2, 1), rosen, method = "Nelder-Mead"); c(nm$par, evals = nm$counts["function"])
```

Robust and derivative-free (R's `optim` default), but slow in high dimensions and offers no standard errors.


## Animation for Nelder-Mead Algorithm

```{shinylive-r}
#| standalone: true
#| viewerHeight: 820

library(shiny)

## ---------------------------------------------------------------- objectives
mk <- function(label, fv, xlim, ylim, start, xstar) {
  force(fv)
  list(label = label, fv = fv, f = function(p) fv(p[1], p[2]),
       xlim = xlim, ylim = ylim, start = start,
       xstar = matrix(xstar, ncol = 2))
}

FNS <- list(
  quad = mk(
    "(a) Well-conditioned quadratic",
    function(x, y) 0.5 * (2 * (x - 1)^2 + 1.6 * (x - 1) * (y - 1) +
                          2 * (y - 1)^2),
    c(-3, 4), c(-3, 4), c(-2, 3), c(1, 1)
  ),
  banana = mk(
    "(b) Rosenbrock banana (narrow curved ridge)",
    function(x, y) (1 - x)^2 + 100 * (y - x^2)^2,
    c(-2, 2), c(-1, 3), c(-1.2, 1), c(1, 1)
  ),
  mix = mk(
    "(c) Highly correlated, two modes",
    function(x, y) {
      rho <- 0.9; den <- 1 - rho^2
      q <- function(u, v) (u^2 - 2 * rho * u * v + v^2) / den
      -log(exp(-0.5 * q(x - 1.2, y - 1.2)) +
           0.6 * exp(-0.5 * q(x + 1.5, y + 1.5) / 1.5) + 1e-12)
    },
    c(-4, 4), c(-4, 4), c(-3, -0.5), c(1.2, 1.2)
  ),
  himmel = mk(
    "(d) Himmelblau (four global minima)",
    function(x, y) (x^2 + y - 11)^2 + (x + y^2 - 7)^2,
    c(-5, 5), c(-5, 5), c(-4, 4),
    rbind(c(3, 2), c(-2.805118, 3.131312),
          c(-3.779310, -3.283186), c(3.584428, -1.848126))
  ),
  beale = mk(
    "(e) Beale (flat plateau, sharp valley)",
    function(x, y) (1.5 - x + x * y)^2 + (2.25 - x + x * y^2)^2 +
                   (2.625 - x + x * y^3)^2,
    c(-4.5, 4.5), c(-4.5, 4.5), c(-1, 1), c(3, 0.5)
  ),
  rastrigin = mk(
    "(f) Rastrigin (many local minima)",
    function(x, y) 20 + x^2 - 10 * cos(2 * pi * x) +
                        y^2 - 10 * cos(2 * pi * y),
    c(-5.12, 5.12), c(-5.12, 5.12), c(-3.1, 4.2), c(0, 0)
  )
)

## ------------------------------------------------------------- Nelder-Mead
## coefficients: reflection a, expansion g, contraction r, shrink s
run_nm <- function(FN, x0, h, a = 1, g = 2, r = 0.5, s = 0.5,
                   maxit = 80, tol = 1e-7) {
  f <- FN$f
  S <- rbind(x0, x0 + c(h, 0), x0 + c(0, h))
  fS <- apply(S, 1, f)
  steps <- list()
  note <- "maximum number of iterations reached"

  for (k in seq_len(maxit)) {
    o <- order(fS); S <- S[o, , drop = FALSE]; fS <- fS[o]
    dia <- max(c(sqrt(sum((S[1, ] - S[2, ])^2)),
                 sqrt(sum((S[1, ] - S[3, ])^2)),
                 sqrt(sum((S[2, ] - S[3, ])^2))))

    if (dia < tol || diff(range(fS)) < 1e-12) {
      steps[[k]] <- list(S = S, fS = fS, dia = dia, op = "converged",
                         cen = colMeans(S[1:2, , drop = FALSE]),
                         xr = NULL, xe = NULL, xc = NULL, Sn = S)
      note <- "converged (simplex collapsed)"
      break
    }

    cen <- colMeans(S[1:2, , drop = FALSE])          # centroid of the best two
    xr <- cen + a * (cen - S[3, ]); fr <- f(xr)
    xe <- NULL; xc <- NULL
    Sn <- S; fn <- fS

    if (fr < fS[1]) {                                 # better than the best
      xe <- cen + g * (xr - cen); fe <- f(xe)
      if (fe < fr) { Sn[3, ] <- xe; fn[3] <- fe; op <- "expansion" }
      else         { Sn[3, ] <- xr; fn[3] <- fr; op <- "reflection" }
    } else if (fr < fS[2]) {                          # middling: accept
      Sn[3, ] <- xr; fn[3] <- fr; op <- "reflection"
    } else {
      if (fr < fS[3]) {                               # outside contraction
        xc <- cen + r * (xr - cen); fc <- f(xc)
        if (fc <= fr) { Sn[3, ] <- xc; fn[3] <- fc; op <- "outside contraction" }
        else op <- "shrink"
      } else {                                        # inside contraction
        xc <- cen + r * (S[3, ] - cen); fc <- f(xc)
        if (fc < fS[3]) { Sn[3, ] <- xc; fn[3] <- fc; op <- "inside contraction" }
        else op <- "shrink"
      }
      if (op == "shrink") {
        Sn[2, ] <- S[1, ] + s * (S[2, ] - S[1, ])
        Sn[3, ] <- S[1, ] + s * (S[3, ] - S[1, ])
        fn[2] <- f(Sn[2, ]); fn[3] <- f(Sn[3, ])
      }
    }

    steps[[k]] <- list(S = S, fS = fS, dia = dia, op = op, cen = cen,
                       xr = xr, xe = xe, xc = xc, Sn = Sn)
    S <- Sn; fS <- fn
  }

  list(steps = steps, n = length(steps), note = note)
}

## ---------------------------- widen the data box to the device aspect ratio
expand_box <- function(xlim, ylim, pin) {
  if (length(pin) != 2 || any(!is.finite(pin)) || any(pin <= 0))
    return(list(xlim = xlim, ylim = ylim))
  rr <- pin[1] / pin[2]
  wd <- diff(xlim); ht <- diff(ylim)
  if (wd / ht < rr) {
    w <- ht * rr; m <- mean(xlim); xlim <- m + c(-0.5, 0.5) * w
  } else {
    hh <- wd / rr; m <- mean(ylim); ylim <- m + c(-0.5, 0.5) * hh
  }
  list(xlim = xlim, ylim = ylim)
}

COL <- c(refl = "#E8820C", exp = "#7B3FBF", con = "#1F6FB4",
         new = "#2E8B57", best = "#2E8B57", worst = "#C81E1E")

## --------------------------------------------------------------------- UI
ui <- fluidPage(
  fluidRow(
    column(
      3,
      selectInput("fn", "Objective function",
                  setNames(names(FNS), sapply(FNS, `[[`, "label"))),
      sliderInput("k", "Iteration k", min = 0, max = 1, value = 0, step = 1),
      sliderInput("h", "Initial simplex size (% of range)",
                  min = 2, max = 40, value = 12, step = 1),
      checkboxInput("showpath", "Keep past simplices", TRUE),
      htmlOutput("info")
    ),
    column(
      2,
      div(
        style = "margin-top:26px;",
        actionButton("play", "Play", class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("step", "Next", class = "btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("back", "Prev", class = "btn-lg",
                     style = "width:100%;"),
        hr(),
        checkboxInput("autoplay", "Autoplay on change", FALSE),
        actionButton("reset", "Reset initial",
                     style = "width:100%; margin-bottom:8px;"),
        helpText("Click inside the contour plot to choose a new initial value.")
      )
    ),
    column(
      7,
      plotOutput("contour", click = "click", height = "540px"),
      plotOutput("conv", height = "190px")
    )
  )
)

## ----------------------------------------------------------------- server
server <- function(input, output, session) {

  FN <- reactive(FNS[[input$fn]])
  start <- reactiveVal(FNS[[1]]$start)
  playing <- reactiveVal(FALSE)

  observeEvent(input$fn, start(FNS[[input$fn]]$start))
  observeEvent(input$reset, start(FN()$start))
  observeEvent(input$click, {
    p <- c(input$click$x, input$click$y)
    if (all(is.finite(p))) start(p)
  })

  path <- reactive({
    req(start())
    FNc <- FN()
    run_nm(FNc, start(), h = input$h / 100 * diff(FNc$xlim))
  })

  observeEvent(path(), {
    updateSliderInput(session, "k", max = max(1, path()$n - 1), value = 0)
    playing(isTRUE(input$autoplay) && path()$n > 1)
  })

  observeEvent(input$play, {
    if (isTRUE(playing())) {
      playing(FALSE)
    } else {
      if (as.integer(input$k) >= path()$n - 1)
        updateSliderInput(session, "k", value = 0)
      playing(TRUE)
    }
  })

  observeEvent(playing(), {
    updateActionButton(session, "play",
                       label = if (isTRUE(playing())) "Pause" else "Play")
  })

  observe({
    if (!isTRUE(playing())) return()
    kmax <- isolate(path()$n) - 1
    k <- isolate(as.integer(input$k))
    if (k >= kmax) { playing(FALSE); return() }
    invalidateLater(900, session)
    updateSliderInput(session, "k", value = k + 1)
  })

  observeEvent(input$step, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = min(as.integer(input$k) + 1, max(0, path()$n - 1)))
  })

  observeEvent(input$back, {
    playing(FALSE)
    updateSliderInput(session, "k", value = max(as.integer(input$k) - 1, 0))
  })

  output$contour <- renderPlot({
    FNc <- FN(); P <- path()
    k <- min(as.integer(input$k), P$n - 1) + 1L
    st <- P$steps[[k]]

    par(mar = c(4, 4, 3, 1))
    plot.new()
    bx <- expand_box(FNc$xlim, FNc$ylim, par("pin"))
    plot.window(xlim = bx$xlim, ylim = bx$ylim, xaxs = "i", yaxs = "i")

    xs <- seq(bx$xlim[1], bx$xlim[2], length.out = 180)
    ys <- seq(bx$ylim[1], bx$ylim[2], length.out = 180)
    z  <- outer(xs, ys, FNc$fv)
    zf <- z[is.finite(z)]
    lv <- unique(quantile(zf, probs = seq(0, 1, length.out = 30)^1.7))
    zr <- matrix(rank(z, na.last = "keep", ties.method = "average"),
                 nrow = length(xs))

    image(xs, ys, zr, add = TRUE,
          col = colorRampPalette(c("#f7fbff", "#9ec4e3"))(128))
    contour(xs, ys, z, levels = lv, add = TRUE,
            col = "grey45", drawlabels = FALSE)
    axis(1); axis(2); box()
    title(main = paste0("Nelder-Mead  |  ", FNc$label),
          xlab = expression(x[1]), ylab = expression(x[2]), cex.main = 1.0)
    mtext(paste0("iteration ", k - 1, ":  ", st$op), side = 3, line = 0.1,
          cex = 1.1, font = 2)

    points(FNc$xstar[, 1], FNc$xstar[, 2], pch = 8, col = "red",
           cex = 1.3, lwd = 2)

    ## past simplices
    if (isTRUE(input$showpath) && k > 1)
      for (j in seq_len(k - 1))
        polygon(P$steps[[j]]$S[, 1], P$steps[[j]]$S[, 2],
                border = "grey55", lty = 3)

    ## current simplex
    S <- st$S
    polygon(S[, 1], S[, 2], border = "black", lwd = 2,
            col = adjustcolor("white", alpha.f = 0.35))
    points(st$cen[1], st$cen[2], pch = 3, col = "black", lwd = 2, cex = 1.1)
    points(S[, 1], S[, 2], pch = 21, cex = 1.6, lwd = 2, bg = "white",
           col = c(COL["best"], "grey30", COL["worst"]))
    text(S[, 1], S[, 2], labels = c("B", "G", "W"), pos = 3, offset = 0.6,
         font = 2, col = c(COL["best"], "grey30", COL["worst"]))

    ## candidate points generated this iteration
    if (!is.null(st$xr)) {
      segments(S[3, 1], S[3, 2], st$xr[1], st$xr[2],
               col = COL["refl"], lwd = 2, lty = 2)
      points(st$xr[1], st$xr[2], pch = 19, col = COL["refl"], cex = 1.5)
    }
    if (!is.null(st$xe)) {
      segments(st$xr[1], st$xr[2], st$xe[1], st$xe[2],
               col = COL["exp"], lwd = 2, lty = 2)
      points(st$xe[1], st$xe[2], pch = 19, col = COL["exp"], cex = 1.5)
    }
    if (!is.null(st$xc))
      points(st$xc[1], st$xc[2], pch = 19, col = COL["con"], cex = 1.5)

    ## the simplex that results
    if (st$op != "converged") {
      polygon(st$Sn[, 1], st$Sn[, 2], border = COL["new"], lwd = 3, lty = 2)
      if (st$op == "shrink")
        arrows(S[2:3, 1], S[2:3, 2], st$Sn[2:3, 1], st$Sn[2:3, 2],
               col = COL["con"], lwd = 2, length = 0.10)
    }

    legend("topleft", cex = 0.85, bg = "#ffffffcc", box.col = NA,
           pch = c(21, 21, 3, 19, 19, 19, NA),
           lty = c(NA, NA, NA, NA, NA, NA, 2),
           lwd = c(2, 2, 2, NA, NA, NA, 3),
           col = c(COL["best"], COL["worst"], "black", COL["refl"],
                   COL["exp"], COL["con"], COL["new"]),
           legend = c("best vertex B", "worst vertex W",
                      "centroid of B and G", "reflection", "expansion",
                      "contraction / shrink", "next simplex"))
  })

  output$conv <- renderPlot({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    fb <- sapply(P$steps, function(s) s$fS[1])
    fw <- sapply(P$steps, function(s) s$fS[3])
    par(mar = c(4, 4.5, 1.5, 1))
    matplot(seq_len(P$n) - 1, cbind(fb, fw), type = "b", pch = 20, lty = 1,
            col = c(COL["best"], COL["worst"]),
            xlab = "iteration k", ylab = "f at simplex vertices")
    abline(v = k - 1, col = "grey60", lty = 2)
    legend("topright", bty = "n", cex = 0.9, lty = 1, pch = 20,
           col = c(COL["best"], COL["worst"]), legend = c("best", "worst"))
  })

  output$info <- renderUI({
    P <- path(); k <- min(as.integer(input$k), P$n - 1) + 1L
    st <- P$steps[[k]]
    fmt <- function(v, d = 4) formatC(v, format = "g", digits = d)
    HTML(paste0(
      "<hr><b>iteration</b> ", k - 1, " of ", P$n - 1, "<br>",
      "<b>operation</b>: ", st$op, "<br><br>",
      "<b>B</b> = (", fmt(st$S[1, 1]), ", ", fmt(st$S[1, 2]), "),  f = ",
      fmt(st$fS[1], 6), "<br>",
      "<b>G</b> = (", fmt(st$S[2, 1]), ", ", fmt(st$S[2, 2]), "),  f = ",
      fmt(st$fS[2], 6), "<br>",
      "<b>W</b> = (", fmt(st$S[3, 1]), ", ", fmt(st$S[3, 2]), "),  f = ",
      fmt(st$fS[3], 6), "<br><br>",
      "<b>simplex diameter</b> = ", fmt(st$dia), "<br>",
      "<b>f range</b> = ", fmt(diff(range(st$fS))),
      "<br><br><i>", P$note, "</i>"
    ))
  })
}

shinyApp(ui, server)
```
# Stochastic Gradient Descent

## Motivation: Large $n$

For iid data the log-likelihood and its gradient are **sums over observations**:
$$
\ell(\theta) = \sum_{i=1}^n \ell_i(\theta), \qquad \nabla\ell(\theta) = \sum_{i=1}^n \nabla\ell_i(\theta).
$$

Every method so far needs the **full** gradient, an $O(np)$ pass through all data, per iteration (and Newton needs $O(np^2)$ for the Hessian). With $n$ in the millions and $p$ in the millions (neural networks), even one full gradient is expensive.

**Idea.** $\nabla\ell_i(\theta)$ for a randomly chosen $i$ is an **unbiased estimate** of the average gradient:
$$
E_i\big[\nabla\ell_i(\theta)\big] = \frac{1}{n}\nabla\ell(\theta).
$$

A noisy but cheap gradient estimate is enough to make progress — a Monte Carlo idea applied to optimization.


## The SGD Algorithm {.smaller}

For $t = 1, 2, \ldots$:

1. Sample an index $i_t$ uniformly from $\{1, \ldots, n\}$ (or shuffle the data and cycle through it; one pass is an **epoch**).

2. Update in the ascent direction of that single observation's log-likelihood:
$$
\theta_{t+1} = \theta_t + \gamma_t\,\nabla\ell_{i_t}(\theta_t).
$$

**Mini-batch SGD** averages the gradient over a batch $B_t$ of $m$ observations ($m = 32$–$512$ typical):
$$
\theta_{t+1} = \theta_t + \gamma_t\,\frac{1}{m}\sum_{i\in B_t}\nabla\ell_i(\theta_t),
$$
reducing the variance by $1/m$ at $m$ times the cost, and using vectorized/GPU arithmetic.

The **learning rate** $\gamma_t$ replaces the line search (a line search would need full-data evaluations).

## Learning Rate and Convergence

Because the gradient is noisy, a constant $\gamma$ makes $\theta_t$ **bounce around** the optimum with variance $\propto \gamma$. Robbins–Monro (1951): convergence to the optimum is guaranteed if
$$
\sum_t \gamma_t = \infty \quad\text{and}\quad \sum_t \gamma_t^2 < \infty, \qquad\text{e.g. } \gamma_t = \frac{\gamma_0}{1 + t/t_0} .
$$

- $\gamma$ too large: divergence or persistent noise. $\gamma$ too small: painfully slow.

- Convergence is only **sublinear**, far slower than Newton *per iteration* — but each iteration costs $O(p)$ instead of $O(np)$, and early progress is fast.
- Common refinements: **momentum** (average past gradients), **Adam** (per-coordinate adaptive rates), **averaging** the iterates $\overline\theta_T = \frac1T\sum_t\theta_t$ (Polyak–Ruppert), which restores the $\sqrt n$-efficiency of the MLE.


## Animation for SGD

```{shinylive-r}
#| label: fig-sgd-animation
#| standalone: true
#| viewerHeight: 830

library(shiny)

## ------------------------------------------------------------------- data
## Two data-generating processes. The teacher network lies inside the
## student's model class; the bumps do not.
TEACH <- list(w = c(4, 3, 5), cc = c(4, 0, -5), v = c(1.0, -1.2, 0.8), b = 0)

g_teacher <- function(x) {
  Hm <- tanh(outer(x, TEACH$w) + rep(TEACH$cc, each = length(x)))
  as.vector(Hm %*% TEACH$v + TEACH$b)
}

g_bumps <- function(x)
  1.2 * exp(-9 * (x + 0.7)^2) - 1.0 * exp(-9 * (x - 0.8)^2)

DGPS <- c("Teacher network (1 - 3 tanh - 1)" = "teacher",
          "Two Gaussian bumps (outside the model class)" = "bumps")

gtrue <- function(x, dgp)
  if (identical(dgp, "bumps")) g_bumps(x) else g_teacher(x)

XG <- seq(-2, 2, length.out = 300)

make_data <- function(n, sigma, seed, dgp) {
  set.seed(500 + seed)
  x <- sort(runif(n, -2, 2))
  list(x = x, y = gtrue(x, dgp) + rnorm(n, 0, sigma), n = n, dgp = dgp)
}

## ------------------------------------------------- network: 1 - H tanh - 1
unpack <- function(p, H)
  list(w = p[1:H], cc = p[H + (1:H)], v = p[2 * H + (1:H)], b = p[3 * H + 1])

fnet <- function(p, x, H) {
  P <- unpack(p, H)
  Hm <- tanh(outer(x, P$w) + rep(P$cc, each = length(x)))
  as.vector(Hm %*% P$v + P$b)
}

gnet <- function(p, x, y, H) {
  n <- length(x); P <- unpack(p, H)
  Hm <- tanh(outer(x, P$w) + rep(P$cc, each = n))
  r <- as.vector(Hm %*% P$v + P$b) - y
  S <- (1 - Hm^2) * outer(r, P$v)
  c(as.vector(colSums(S * x)) / n,        # dw
    as.vector(colSums(S)) / n,            # dc
    as.vector(crossprod(Hm, r)) / n,      # dv
    mean(r))                              # db
}

mse <- function(p, H, x, y) 0.5 * mean((fnet(p, x, H) - y)^2)

init_p <- function(H, seed) {
  set.seed(1000 + seed)
  c(rnorm(H, 0, 2), rnorm(H, 0, 1.5), rnorm(H, 0, 0.7), 0)
}

## ---------------------------------------------------------------- training
## m = batch size; m >= n is full-batch gradient descent.
train <- function(p0, H, m, a, nepoch, seed, D) {
  n <- D$n
  m <- max(1, min(m, n))
  B <- ceiling(n / m)                              # updates per epoch
  nup <- B * nepoch
  P <- matrix(NA_real_, nup + 1, length(p0))
  L <- rep(NA_real_, nup + 1); EP <- rep(NA_real_, nup + 1)
  IDX <- vector("list", nup + 1)
  set.seed(seed)
  p <- p0; k <- 0; bad <- FALSE
  for (e in seq_len(nepoch)) {
    perm <- if (B == 1) seq_len(n) else sample.int(n)
    grp <- split(perm, ceiling(seq_len(n) / m))    # last batch may be smaller
    for (b in seq_len(B)) {
      k <- k + 1
      idx <- grp[[b]]
      P[k, ] <- p; IDX[[k]] <- idx
      L[k] <- mse(p, H, D$x, D$y); EP[k] <- (k - 1) / B
      p <- p - a * gnet(p, D$x[idx], D$y[idx], H)
      if (any(!is.finite(p))) { bad <- TRUE; break }
    }
    if (bad) break
  }
  if (!bad) {
    k <- k + 1
    P[k, ] <- p; IDX[[k]] <- seq_len(n)
    L[k] <- mse(p, H, D$x, D$y); EP[k] <- (k - 1) / B
  }
  n_rec <- max(1, k)
  list(P = P[1:n_rec, , drop = FALSE], L = L[1:n_rec], EP = EP[1:n_rec],
       IDX = IDX[1:n_rec], n = n_rec, B = B, m = m, diverged = bad)
}

## ------------------------------------------------------------ weight plot
## Diverging ramp: blue = negative, red = positive, magnitude by saturation.
DIV <- colorRampPalette(c("#08306B", "#2171B5", "#6BAED6", "#DEEBF7",
                          "#F7F7F7",
                          "#FEE0D2", "#FC9272", "#CB181D", "#67000D"))(201)

## compander: maps v in [-wmax, wmax] to [-1, 1], boosting small values.
tone <- function(v, wmax, k) {
  u <- pmax(-1, pmin(1, v / wmax))
  sign(u) * log1p(k * abs(u)) / log1p(k)
}

wcol <- function(v, wmax, k) DIV[round((tone(v, wmax, k) + 1) * 100) + 1]
wlwd <- function(v, wmax, k) 0.8 + 6 * abs(tone(v, wmax, k))

draw_net <- function(p, H, wmax, k, main) {
  P <- unpack(p, H)
  yh <- if (H == 1) 0.5 else seq(0.92, 0.10, length.out = H)
  ix <- 0.05; hx <- 0.55; ox <- 1.05
  nc <- max(1.4, min(4.4, 36 / H))
  plot.new(); plot.window(c(-0.05, 1.2), c(-0.10, 1.02))
  title(main = main, cex.main = 1.05)
  for (j in 1:H)
    segments(ix, 0.5, hx, yh[j], col = wcol(P$w[j], wmax, k),
             lwd = wlwd(P$w[j], wmax, k), lend = 1)
  for (j in 1:H)
    segments(hx, yh[j], ox, 0.5, col = wcol(P$v[j], wmax, k),
             lwd = wlwd(P$v[j], wmax, k), lend = 1)
  points(ix, 0.5, pch = 21, cex = 3.2, bg = "white", lwd = 1.6)
  text(ix, 0.5, "x", font = 2)
  points(rep(hx, H), yh, pch = 21, cex = nc, lwd = 1.4,
         bg = wcol(P$cc, wmax, k))
  if (H <= 10) text(rep(hx, H), yh, 1:H, cex = 0.65)
  points(ox, 0.5, pch = 21, cex = 3.4, lwd = 1.6, bg = wcol(P$b, wmax, k))
  text(ox, 0.5, "y", font = 2)

  ## signed colour bar
  xs <- seq(0.12, 0.98, length.out = 201)
  rect(xs, -0.09, xs + diff(xs)[1], -0.035, col = DIV, border = NA)
  lab <- formatC(wmax, format = "g", digits = 2)
  text(c(0.12, 0.55, 0.98), -0.005,
       c(paste0("-", lab), "0", paste0("+", lab)), cex = 0.72)
}

## --------------------------------------------------------------------- UI
ui <- fluidPage(
  fluidRow(
    ## ---- column 1: training controls -----------------------------------
    column(
      3,
      selectInput("dgp", "True curve", choices = DGPS, selected = "teacher"),
      sliderInput("H", "Hidden neurons", min = 2, max = 16, value = 8,
                  step = 1),
      sliderInput("m", "Batch size m", min = 1, max = 400, value = 20,
                  step = 1),
      sliderInput("a", "Learning rate", min = 0.01, max = 0.60, value = 0.20,
                  step = 0.01),
      sliderInput("nep", "Epochs", min = 10, max = 300, value = 100, step = 10),
      sliderInput("ctr", "Weight colour contrast", min = 1, max = 40,
                  value = 12, step = 1),
      checkboxInput("showgd", "Overlay the full-batch run", TRUE)
    ),
    ## ---- column 2: data generation, then playback ----------------------
    column(
      2,
      div(
        style = "margin-top:10px;",
        tags$h5(tags$b("Data")),
        sliderInput("n", "Sample size n", min = 40, max = 400, value = 200,
                    step = 20),
        sliderInput("sigma", "Noise SD", min = 0.01, max = 0.40, value = 0.08,
                    step = 0.01),
        actionButton("newdata", "Regenerate a dataset",
                     class = "btn-warning",
                     style = "width:100%; margin-bottom:6px;"),

        hr(style = "border-top:2px solid #999; margin:14px 0;"),

        tags$h5(tags$b("Training")),
        actionButton("play", "Play", class = "btn-primary btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("step", "Next", class = "btn-lg",
                     style = "width:100%; margin-bottom:10px;"),
        actionButton("back", "Prev", class = "btn-lg",
                     style = "width:100%;"),
        checkboxInput("byepoch", "Advance a whole epoch", TRUE),
        checkboxInput("autoplay", "Autoplay on change", FALSE),
        actionButton("newinit", "New initial weights",
                     style = "width:100%; margin-bottom:8px;"),
        actionButton("newseed", "New batch sequence",
                     style = "width:100%; margin-bottom:8px;"),
        helpText("Both runs share the data, the initial weights and the",
                 "number of epochs; only m differs.")
      )
    ),
    ## ---- column 3: plots -----------------------------------------------
    column(
      7,
      plotOutput("nets", height = "300px"),
      plotOutput("fit", height = "340px"),
      sliderInput("k", "Update k", min = 0, max = 1, value = 0, step = 1,
                  width = "900px")
    )
  )
)

## ----------------------------------------------------------------- server
server <- function(input, output, session) {

  MAXUP <- 4000
  dseed <- reactiveVal(1); wseed <- reactiveVal(1); bseed <- reactiveVal(1)
  playing <- reactiveVal(FALSE)

  observeEvent(input$newdata, dseed(dseed() + 1))
  observeEvent(input$newinit, wseed(wseed() + 1))
  observeEvent(input$newseed, bseed(bseed() + 1))

  observeEvent(input$n, {
    updateSliderInput(session, "m", max = input$n,
                      value = min(input$m, input$n))
  })

  DAT <- reactive(make_data(input$n, input$sigma, dseed(), input$dgp))

  runs <- reactive({
    D <- DAT(); H <- input$H
    m <- max(1, min(input$m, D$n))
    B <- ceiling(D$n / m)
    ne <- max(1, min(input$nep, floor(MAXUP / B)))
    p0 <- init_p(H, wseed())
    S <- train(p0, H, m, input$a, ne, bseed(), D)
    G <- train(p0, H, D$n, input$a, ne, bseed(), D)
    list(S = S, G = G, H = H, B = B, m = m, D = D, ne = ne,
         capped = ne < input$nep,
         wmax = max(0.2, quantile(abs(c(S$P, G$P)), 0.97, na.rm = TRUE)))
  })

  stride <- reactive(if (isTRUE(input$byepoch)) runs()$B else 1L)

  observeEvent(runs(), {
    updateSliderInput(session, "k", max = max(1, runs()$S$n - 1), value = 0)
    playing(isTRUE(input$autoplay) && runs()$S$n > 1)
  })

  observeEvent(input$play, {
    if (isTRUE(playing())) playing(FALSE)
    else {
      if (as.integer(input$k) >= runs()$S$n - 1)
        updateSliderInput(session, "k", value = 0)
      playing(TRUE)
    }
  })
  observeEvent(playing(), {
    updateActionButton(session, "play",
                       label = if (isTRUE(playing())) "Pause" else "Play")
  })
  observe({
    if (!isTRUE(playing())) return()
    kmax <- isolate(runs()$S$n) - 1
    k <- isolate(as.integer(input$k)); st <- isolate(stride())
    if (k >= kmax) { playing(FALSE); return() }
    invalidateLater(250, session)
    updateSliderInput(session, "k", value = min(k + st, kmax))
  })
  observeEvent(input$step, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = min(as.integer(input$k) + stride(),
                                  max(0, runs()$S$n - 1)))
  })
  observeEvent(input$back, {
    playing(FALSE)
    updateSliderInput(session, "k",
                      value = max(as.integer(input$k) - stride(), 0))
  })

  cur <- reactive({
    R <- runs()
    k <- min(as.integer(input$k), R$S$n - 1) + 1L
    ep <- R$S$EP[k]
    kg <- max(1, min(R$G$n, floor(ep) + 1L))
    list(R = R, k = k, kg = kg, ep = ep)
  })

  output$nets <- renderPlot({
    C <- cur(); R <- C$R
    msz <- length(R$S$IDX[[C$k]])
    par(mfrow = c(1, if (isTRUE(input$showgd)) 2 else 1), mar = c(1, 1, 3, 1))
    draw_net(R$S$P[C$k, ], R$H, R$wmax, input$ctr,
             sprintf("m = %d (B = %d),  epoch %.2f", msz, R$B, C$ep))
    if (isTRUE(input$showgd))
      draw_net(R$G$P[C$kg, ], R$H, R$wmax, input$ctr,
               sprintf("m = %d (full batch),  epoch %d", R$D$n, C$kg - 1L))
  })

  output$fit <- renderPlot({
    C <- cur(); R <- C$R; D <- R$D
    par(mfrow = c(1, 2), mar = c(4, 4.2, 2.5, 1))

    idx <- R$S$IDX[[C$k]]
    plot(D$x, D$y, pch = 20, col = "grey75", cex = 0.9,
         xlab = "x", ylab = "y", main = "fitted curve and current batch")
    points(D$x[idx], D$y[idx], pch = 19, col = "#E8820C", cex = 1.2)
    lines(XG, gtrue(XG, D$dgp), col = "red", lty = 2, lwd = 2)
    lines(XG, fnet(R$S$P[C$k, ], XG, R$H), col = "#2E8B57", lwd = 2.5)
    if (isTRUE(input$showgd))
      lines(XG, fnet(R$G$P[C$kg, ], XG, R$H), col = "#7B3FBF", lwd = 2.5)
    legend("topright", bty = "n", cex = 0.85,
           lwd = c(2, 2.5, 2.5), lty = c(2, 1, 1),
           col = c("red", "#2E8B57", "#7B3FBF"),
           legend = c(if (identical(D$dgp, "bumps")) "true curve (bumps)"
                      else "teacher network",
                      paste0("m = ", R$m, " fit"), "full-batch fit"))

    yl <- range(c(R$S$L, if (isTRUE(input$showgd)) R$G$L), na.rm = TRUE)
    plot(R$S$EP, R$S$L, type = "l", log = "y", col = "#2E8B57", lwd = 2,
         ylim = pmax(yl, 1e-7), xlab = "epoch (one pass over all n)",
         ylab = "training loss (all n)", main = "loss per epoch")
    if (isTRUE(input$showgd))
      lines(R$G$EP, R$G$L, col = "#7B3FBF", lwd = 2)
    abline(h = 0.5 * input$sigma^2, col = "grey55", lty = 3)
    points(C$ep, R$S$L[C$k], pch = 19, col = "#E8820C", cex = 1.5)
  })

  output$info <- renderUI({
    C <- cur(); R <- C$R; H <- R$H
    P <- unpack(R$S$P[C$k, ], H)
    fmt <- function(v, d = 3) formatC(v, format = "f", digits = d)
    rows <- paste0(
      "<tr><td>h", 1:H, "</td><td align='right'>", fmt(P$w),
      "</td><td align='right'>", fmt(P$cc),
      "</td><td align='right'>", fmt(P$v), "</td></tr>", collapse = "")
    HTML(paste0(
      "<hr><b>update</b> ", C$k - 1, " of ", R$S$n - 1,
      "  (epoch ", fmt(C$ep, 2), ")<br>",
      "<b>batch size m</b> = ", length(R$S$IDX[[C$k]]), " of ", R$D$n, "<br>",
      "<b>updates per epoch</b> = ", R$B, "<br>",
      "<b>parameters</b> = ", 3 * H + 1, "<br><br>",
      "<b>loss, m = ", R$m, "</b> = ", fmt(R$S$L[C$k], 5), "<br>",
      "<b>loss, full batch</b> = ", fmt(R$G$L[C$kg], 5), "<br>",
      "<b>noise floor</b> &asymp; ", fmt(0.5 * input$sigma^2, 5), "<br>",
      if (isTRUE(R$S$diverged))
        "<b style='color:#C81E1E'>run diverged</b><br>" else "",
      if (isTRUE(R$capped))
        paste0("<i style='color:#C81E1E'>epochs capped at ", R$ne,
               "</i><br>") else "",
      "<br><table style='font-size:90%'>",
      "<tr><th>unit</th><th>w</th><th>bias</th><th>v</th></tr>",
      rows, "</table>",
      "<br><i>output bias = ", fmt(P$b), "</i>"
    ))
  })
}

shinyApp(ui, server)
```


## Example: SGD for Logistic Regression in R

```{r}
sgd <- function(X, y, b0, gamma0 = 0.05, t0 = 200, epochs = 5, batch = 10) {
  n <- nrow(X); b <- b0; t <- 0; path <- b0
  for (e in 1:epochs) {
    idx <- sample(n)
    for (s in seq(1, n, by = batch)) {
      i <- idx[s:min(s + batch - 1, n)]; t <- t + 1
      g <- crossprod(X[i, , drop = FALSE], y[i] - plogis(X[i, , drop = FALSE] %*% b)) / length(i)
      b <- b + gamma0 / (1 + t / t0) * g; path <- rbind(path, drop(b)) } }
  list(est = drop(b), path = path)
}
sg <- sgd(X, y, rep(0, 3))
rbind(sgd = sg$est, newton = fit$est, truth = beta_true)
```

## SGD Trajectory vs. Newton

```{r}
#| echo: false
m <- 10; per_epoch <- nrow(X) / m
xs <- 0:(nrow(sg$path) - 1); xn <- (0:fit$iter) * per_epoch
matplot(xs, sg$path[, 2:3], type = "l", lty = 1, col = c(2, 4), ylim = range(sg$path[, 2:3], fit$path[, 2:3]),
        xlab = "mini-batch updates (one epoch = one full-data pass)", ylab = "coefficient",
        main = expression(paste("SGD vs. Newton: ", beta[1], " (red), ", beta[2], " (blue)")))
matlines(xn, fit$path[, 2:3], type = "b", lty = 2, pch = 19, col = c(2, 4), lwd = 2)
abline(h = fit$est[2:3], col = c(2, 4), lty = 3); abline(v = seq(0, max(xs), per_epoch), col = "grey85")
legend("right", c("SGD (mini-batch)", "Newton (per full pass)", "MLE"), lty = c(1, 2, 3), pch = c(NA, 19, NA), bty = "n")
```

- Newton needed `r fit$iter` full-data passes; SGD made `r nrow(sg$path) - 1` cheap updates over 5 epochs and lands near the MLE with residual noise.

- Per data pass, Newton is far more efficient here ($n = 500$, $p = 3$); SGD earns its keep when one full gradient — or the $O(p^3)$ Hessian solve — is prohibitively expensive.

## Summary {.smaller}

| | needs | strength | weakness |
|:--|:--|:--|:--|
| bisection / golden section | bracket | always converges | 1-D only, linear |
| Newton–Raphson | $\nabla\ell$, $\nabla^2\ell$ | quadratic convergence, SEs from Hessian | can diverge; $O(p^3)$ |
| Fisher scoring / IRLS | $\nabla\ell$, $I(\theta)$ | positive-definite steps; GLMs | needs expectations |
| Nelder–Mead | $\ell$ | derivative-free | slow, small $p$ |
| gradient descent / CG / BFGS | $\nabla\ell$ | scales to large $p$ | line search; ill-conditioning |
| SGD | $\nabla\ell_i$ | scales to large $n$ and $p$ | noisy, tuning of $\gamma_t$ |

The log-likelihood is approximately quadratic near the MLE (asymptotic normality); every method above exploits that, and its curvature at the optimum delivers the standard errors.
